# Antibody Data (ADAB)

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Anti-drug antibody (ADA) analyses are performed to assess the
immunogenicity of biologic drugs. The `ADAB` dataset contains
antibody titre and associated analysis variables derived from the `PC`
or `IS` (Immunogenicity Specimen Assessments) SDTM domains.

Common ADA parameters include:

-   Titre of anti-drug antibodies (ADA titre)
-   Treatment-emergent ADA status (`TEADAFL`)
-   Persistent ADA status (`PERSAFL`)
-   Neutralising antibody status (`NABFL`)

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(tibble)
library(lubridate)
library(pharmaversesdtm)
```

## Programming Workflow

1.  [Read in Data](#readdata)
2.  [Derive Parameters](#params)
3.  [Derive ADA Flags](#flags)
4.  [Add ADSL Variables](#adsl_vars)
5.  [Derive Analysis Sequence Number](#aseq)

## Read in Data {#readdata}

```{r load-data}
# Using PC domain as a proxy for antibody titre data
# (pharmaversesdtm does not include IS domain)
pc   <- pharmaversesdtm::pc   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl

# Filter to relevant subjects
ab_subjects <- unique(pc$USUBJID)[1:4]
pc_sub  <- pc   |> filter(USUBJID %in% ab_subjects)
adsl_sub <- adsl |> filter(USUBJID %in% ab_subjects)
```

## Derive Parameters {#params}

```{r params}
adab <- pc_sub |>
  derive_vars_dtm(
    dtc             = PCDTC,
    new_vars_prefix = "A",
    flag_imputation = "none"
  ) |>
  mutate(
    PARAMCD = PCTESTCD,
    PARAM   = PCTEST,
    AVAL    = PCSTRESN,
    AVALC   = PCSTRESC,
    AVALU   = PCSTRESU,
    AVISIT  = VISIT,
    AVISITN = VISITNUM
  ) |>
  derive_vars_merged(
    dataset_add = adsl_sub,
    new_vars    = exprs(TRTSDT),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adab |>
  select(USUBJID, PARAMCD, PARAM, AVAL, AVALC, ADTM, AVISIT) |>
  head(8)
```

## Derive ADA Flags {#flags}

### Baseline ADA Status

Identify subjects who are ADA positive at baseline (pre-treatment):

```{r baseline}
adab <- adab |>
  mutate(
    # Flag pre-dose / screening records
    PREABAFL = if_else(
      !is.na(TRTSDT) & as.Date(ADTM) <= TRTSDT & !is.na(AVAL),
      "Y",
      NA_character_
    )
  )

adab |>
  select(USUBJID, ADTM, TRTSDT, PARAMCD, AVAL, PREABAFL) |>
  head(5)
```

### Treatment-Emergent ADA (TEADA) Flag

A subject has treatment-emergent ADA if they are ADA-negative at
baseline but ADA-positive post-dose, or if their post-dose titre is
≥ 4-fold greater than their baseline titre:

```{r teada}
adab <- adab |>
  # Derive baseline titre
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADTM),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & as.Date(ADTM) <= TRTSDT
  ) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  ) |>
  mutate(
    # Simplified TEADA logic (treatment-emergent = post-dose positive)
    TEADAFL = if_else(
      as.Date(ADTM) > TRTSDT & !is.na(AVAL) & AVAL > 0,
      "Y",
      NA_character_
    )
  )

adab |>
  filter(!is.na(TEADAFL)) |>
  select(USUBJID, ADTM, TRTSDT, PARAMCD, AVAL, BASE, TEADAFL) |>
  head(5)
```

### Subject-Level ADA Summary Flags

Subject-level flags are merged into `ADSL`:

```{r subject-flags}
adab_subj <- adab |>
  filter(TEADAFL == "Y") |>
  distinct(STUDYID, USUBJID) |>
  mutate(ADAAFLG = "Y")

adsl_sub <- adsl_sub |>
  left_join(adab_subj, by = c("STUDYID", "USUBJID"))

adsl_sub |>
  select(USUBJID, ADAAFLG) |>
  head(5)
```

## Add ADSL Variables {#adsl_vars}

```{r adsl-vars}
adab <- adab |>
  derive_vars_merged(
    dataset_add = adsl_sub,
    new_vars    = exprs(TRT01A, TRT01P, AGE, SEX),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adab |>
  select(USUBJID, PARAMCD, AVAL, TRT01A, AGE, SEX) |>
  head(5)
```

## Derive Analysis Sequence Number {#aseq}

```{r aseq}
adab <- adab |>
  derive_var_obs_number(
    by_vars    = exprs(STUDYID, USUBJID),
    order      = exprs(PARAMCD, ADTM),
    new_var    = ASEQ,
    check_type = "error"
  )

adab |>
  select(USUBJID, PARAMCD, ADTM, ASEQ) |>
  head(5)
```

## Example Script

An ADAB template script is available in the `{admiral}` package:

```r
use_ad_template("ADAB")
```

## See Also

-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [PK NCA](pk_adnca.qmd)
-   [Date/Time Imputation](imputation.qmd)
