# Creating ADSL

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter describes creating an `ADSL` ADaM dataset. Examples are presented
and tested using `DM`, `EX`, `AE`, `LB`, and `DS` SDTM domains. However,
other domains could be used.

::: callout-note
All examples assume CDISC SDTM and/or ADaM format as input unless otherwise
specified.
:::

## Programming Flow

1.  [Read in Data](#readdata)
2.  [Derive Period, Subperiod, and Phase Variables](#periodvars)
3.  [Derive Treatment Variables](#treatmentvar)
4.  [Derive/Impute Numeric Treatment Date/Time and Duration](#trtdatetime)
5.  [Derive Disposition Variables](#disposition)
6.  [Derive Birth Date and Analysis Age](#aage)
7.  [Derive Death Variables](#death)
8.  [Derive Last Known Date Alive](#lstalvdt)
9.  [Derive Groupings and Populations](#groupings)
10. [Derive Other Variables](#other)
11. [Add Labels and Attributes](#attributes)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADSL` should be read into
the environment. This will be a company-specific process. Some of the data
frames needed may be `DM`, `EX`, `DS`, `AE`, and `LB`.

For example purposes, the CDISC Pilot SDTM datasets — which are included in
`{pharmaversesdtm}` — are used.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
library(stringr)
```

```{r load-data}
dm <- pharmaversesdtm::dm |> convert_blanks_to_na()
ds <- pharmaversesdtm::ds |> convert_blanks_to_na()
ex <- pharmaversesdtm::ex |> convert_blanks_to_na()
ae <- pharmaversesdtm::ae |> convert_blanks_to_na()
lb <- pharmaversesdtm::lb |> convert_blanks_to_na()
```

The `DM` domain is used as the basis for `ADSL`:

```{r adsl-base}
adsl <- dm |> select(-DOMAIN)
adsl |>
  select(USUBJID, RFSTDTC, COUNTRY, AGE, SEX, RACE, ETHNIC, ARM, ACTARM) |>
  head(5)
```

## Derive Period, Subperiod, and Phase Variables {#periodvars}

See the [Visit and Period Variables](visits_periods.qmd#periods_adsl) chapter
for more information.

If the variables are not derived based on a period reference dataset, they may
be derived at a later point of the flow. For example, phases like "Treatment
Phase" and "Follow up" could be derived based on treatment start and end date.

## Derive Treatment Variables {#treatmentvar}

The mapping of the treatment variables is left to the ADaM programmer. An
example mapping for a study without periods may be:

```{r treatment-vars}
adsl <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM)
```

For studies with periods, see the [Visit and Period Variables](visits_periods.qmd#treatment_adsl)
chapter.

## Derive/Impute Numeric Treatment Date/Time and Duration {#trtdatetime}

The function `derive_vars_merged()` can be used to derive the treatment start
and end date/times using the `ex` domain. A pre-processing step for `ex` is
required to convert `EXSTDTC` and `EXENDTC` to datetime variables and impute
missing date or time components. Conversion and imputation is done by
`derive_vars_dtm()`.

```{r treatment-datetime}
# Impute start and end time of exposure to first and last respectively
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST"
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    time_imputation = "last"
  )

adsl <- adsl |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = (EXDOSE > 0 |
      (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = (EXDOSE > 0 |
      (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM, TRTETMF = EXENTMF),
    order       = exprs(EXENDTM, EXSEQ),
    mode        = "last",
    by_vars     = exprs(STUDYID, USUBJID)
  )
```

The datetime variables returned can be converted to dates using
`derive_vars_dtm_to_dt()`.

```{r datetime-to-date}
adsl <- adsl |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))
```

Now that `TRTSDT` and `TRTEDT` are derived, `derive_var_trtdurd()` can be used
to calculate the Treatment Duration (`TRTDURD`).

```{r trtdurd}
adsl <- adsl |>
  derive_var_trtdurd()

adsl |>
  select(USUBJID, RFSTDTC, TRTSDTM, TRTSDT, TRTEDTM, TRTEDT, TRTDURD) |>
  head(5)
```

## Derive Disposition Variables {#disposition}

### Disposition Dates (e.g. `EOSDT`) {#disposition_date}

The functions `derive_vars_dt()` and `derive_vars_merged()` can be used to
derive a disposition date. First, the character disposition date
(`DS.DSSTDTC`) is converted to a numeric date (`EOSDT`) using
`derive_vars_dt()`.

```{r eosdt}
adsl <- adsl |>
  derive_vars_merged(
    dataset_add   = ds,
    filter_add    = DSDECOD == "STUDY DISCONTINUATION",
    new_vars      = exprs(EOSSTT = DSTERM, EOSDTC = DSSTDTC),
    by_vars       = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(
    new_vars_prefix = "EOS",
    dtc             = EOSDTC
  )

adsl |>
  select(USUBJID, EOSSTT, EOSDT) |>
  head(5)
```

### Randomization Date (`RANDDT`) {#randomization_date}

```{r randdt}
adsl <- adsl |>
  derive_vars_merged(
    dataset_add = ds,
    filter_add  = DSDECOD == "RANDOMIZED",
    new_vars    = exprs(RANDDTC = DSSTDTC),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(
    new_vars_prefix = "RAND",
    dtc             = RANDDTC
  )

adsl |>
  select(USUBJID, RANDDT) |>
  head(5)
```

## Derive Birth Date and Analysis Age {#aage}

The function `derive_vars_dt()` can be used to derive the birth date from
`BRTHDTC`, and then `derive_vars_age()` can be used to calculate age.

```{r age}
adsl <- adsl |>
  derive_vars_dt(
    new_vars_prefix = "BRTH",
    dtc             = BRTHDTC
  )

adsl |>
  select(USUBJID, BRTHDTC, BRTHDT) |>
  head(5)
```

## Derive Death Variables {#death}

### Death Date (`DTHDT`) {#death_date}

```{r dthdt}
adsl <- adsl |>
  derive_vars_merged(
    dataset_add = ds,
    filter_add  = DSDECOD == "DEATH",
    new_vars    = exprs(DTHDTC = DSSTDTC),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(
    new_vars_prefix = "DTH",
    dtc             = DTHDTC
  )

adsl |>
  select(USUBJID, DTHDTC, DTHDT) |>
  head(5)
```

## Derive Groupings and Populations {#groupings}

### Grouping (e.g. `AGEGR1`, `REGION1`) {#groupings_ex}

Groupings can be derived using simple `dplyr::mutate()` calls:

```{r groupings}
adsl <- adsl |>
  mutate(
    AGEGR1 = case_when(
      AGE < 65             ~ "<65",
      AGE >= 65 & AGE < 80 ~ "65-80",
      AGE >= 80            ~ ">80"
    ),
    AGEGR1N = case_when(
      AGE < 65             ~ 1,
      AGE >= 65 & AGE < 80 ~ 2,
      AGE >= 80            ~ 3
    )
  )

adsl |>
  select(USUBJID, AGE, AGEGR1, AGEGR1N) |>
  head(5)
```

### Population Flags (e.g. `SAFFL`) {#popflag}

Population flags can be derived using `derive_var_merged_exist_flag()`:

```{r saffl}
adsl <- adsl |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

adsl |>
  select(USUBJID, SAFFL) |>
  head(5)
```

## Add Labels and Attributes {#attributes}

Note that attributes may not be preserved in some cases after processing with
`{admiral}`. The recommended approach is to apply variable labels and other
metadata as a final step in your data derivation process using packages like:

-   [metacore](https://atorus-research.github.io/metacore/): establish a common
    foundation for the use of metadata within an R session.
-   [metatools](https://pharmaverse.github.io/metatools/): enable the use of
    metacore objects to build datasets or enhance existing datasets.
-   [xportr](https://atorus-research.github.io/xportr/): associate all metadata
    information to a local R data frame and convert to a transport v5 file.

## Example Script

An ADSL template script is available in the `{admiral}` package:

```r
use_ad_template("ADSL")
```

## See Also

-   [Visit and Period Variables](visits_periods.qmd)
-   [Date/Time Imputation](imputation.qmd)
-   [pharmaverse E2E Example](https://pharmaverse.github.io/examples/adam/adsl)
