# Creating a BDS Findings ADaM

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter describes creating a BDS (Basic Data Structure) Findings ADaM.
Examples are currently presented and tested using the `VS` (Vital Signs) SDTM
domain to create `ADVS`. However, the workflow applies to all BDS Findings
ADaMs such as `ADLB`, `ADEG`, etc.

::: callout-note
All examples assume CDISC SDTM and/or ADaM format as input unless otherwise
specified.
:::

## Programming Workflow

1.  [Read in Data](#readdata)
2.  [Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1`](#paramcd)
3.  [Derive Analysis Value (`AVAL`, `AVALC`)](#aval)
4.  [Derive Additional Parameters](#params)
5.  [Derive Timing Variables (`ADT`, `ADY`, `AVISIT`, etc.)](#timing)
6.  [Derive Baseline (`ABLFL`, `BASE`, `BASEC`)](#baseline)
7.  [Derive Change from Baseline (`CHG`, `PCHG`)](#chg)
8.  [Derive Reference Range Indicators (`ANRLO`, `ANRHI`, etc.)](#anr)
9.  [Derive Normal Range Flags (`ANL01FL`, etc.)](#anrfl)
10. [Derive Analysis Flags (e.g. `ANL01FL`)](#anlfl)
11. [Derive Treatment and Period Variables](#trt)
12. [Add ADSL Variables](#adsl_vars)
13. [Derive Analysis Sequence Number (`ASEQ`)](#aseq)
14. [Add Labels and Attributes](#attributes)

## Read in Data {#readdata}

For example purposes, the CDISC Pilot SDTM datasets — which are included in
`{pharmaversesdtm}` — are used.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
```

```{r load-data}
vs   <- pharmaversesdtm::vs   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl
```

```{r filter-data}
vs <- vs |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:6])
```

## Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1` {#paramcd}

The function `derive_vars_merged()` can be used to assign parameter-level
variables from a look-up table. Here we use a simple inline look-up table for
the vital signs parameters.

```{r paramcd}
param_lookup <- tribble(
  ~VSTESTCD, ~PARAMCD,    ~PARAM,                             ~PARAMN, ~PARCAT1,
  "SYSBP",   "SYSBP",     "Systolic Blood Pressure (mmHg)",   1,       "VITAL SIGNS",
  "DIABP",   "DIABP",     "Diastolic Blood Pressure (mmHg)",  2,       "VITAL SIGNS",
  "PULSE",   "PULSE",     "Pulse Rate (beats/min)",            3,       "VITAL SIGNS",
  "WEIGHT",  "WEIGHT",    "Weight (kg)",                       4,       "VITAL SIGNS",
  "HEIGHT",  "HEIGHT",    "Height (cm)",                       5,       "VITAL SIGNS",
  "TEMP",    "TEMP",      "Temperature (°C)",                  6,       "VITAL SIGNS",
  "RESP",    "RESP",      "Respiratory Rate (breaths/min)",    7,       "VITAL SIGNS"
)

advs <- vs |>
  derive_vars_merged(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD)
  )

advs |>
  select(USUBJID, VSTESTCD, PARAMCD, PARAM, PARAMN, PARCAT1) |>
  head(5)
```

## Derive Analysis Value {#aval}

For findings ADaMs, `AVAL` usually comes directly from the numeric standard
result (`VSSTRESN`), and `AVALC` from the character standard result
(`VSSTRESC`).

```{r aval}
advs <- advs |>
  mutate(
    AVAL  = VSSTRESN,
    AVALC = VSSTRESC
  )
```

## Derive Additional Parameters {#params}

Additional parameters can be derived using `derive_param_computed()`. For
example, Pulse Pressure (`PULPR`) can be derived from Systolic and Diastolic
Blood Pressure:

```{r additional-params}
advs <- advs |>
  derive_param_computed(
    by_vars       = exprs(USUBJID, VISIT, VISITNUM, VSTPT),
    parameters    = c("SYSBP", "DIABP"),
    set_values_to = exprs(
      AVAL    = AVAL.SYSBP - AVAL.DIABP,
      PARAMCD = "PULPR",
      PARAM   = "Pulse Pressure (mmHg)",
      PARAMN  = 8L,
      PARCAT1 = "VITAL SIGNS"
    )
  )

advs |>
  filter(PARAMCD == "PULPR") |>
  select(USUBJID, VISIT, PARAMCD, PARAM, AVAL) |>
  head(5)
```

## Derive Timing Variables {#timing}

Timing variables include `ADT`, `ADY`, `AVISIT`, `AVISITN`, and `ATPT`.

```{r timing}
advs <- advs |>
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc             = VSDTC
  ) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)
  ) |>
  mutate(
    AVISIT  = if_else(str_detect(VISIT, "SCREEN"), "Baseline", VISIT),
    AVISITN = VISITNUM,
    ATPT    = VSTPT,
    ATPTN   = VSTPTNUM
  )
```

::: callout-tip
Use the `{admiral}` function `derive_vars_merged()` with a visits look-up
table if your `AVISIT` assignment is complex. See [Visit and Period
Variables](visits_periods.qmd) for details.
:::

```{r show-timing}
advs |>
  select(USUBJID, ADT, ADY, AVISIT, AVISITN, ATPT) |>
  head(5)
```

## Derive Baseline {#baseline}

The baseline flag `ABLFL` marks the last non-missing record before (or on)
the treatment start date:

```{r ablfl}
advs <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, VISITNUM),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )
```

Baseline value (`BASE`) and baseline character value (`BASEC`) can then be
derived:

```{r base}
advs <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  ) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVALC,
    new_var    = BASEC
  )

advs |>
  select(USUBJID, PARAMCD, ADT, AVAL, ABLFL, BASE) |>
  head(8)
```

## Derive Change from Baseline {#chg}

```{r chg}
advs <- advs |>
  derive_var_chg() |>
  derive_var_pchg()

advs |>
  select(USUBJID, PARAMCD, ADT, AVAL, BASE, CHG, PCHG) |>
  head(8)
```

## Derive Reference Range Indicators {#anr}

The reference range variables (`ANRLO`, `ANRHI`) can be derived from SDTM:

```{r anr}
advs <- advs |>
  mutate(
    ANRLO = VSSTNRLO,
    ANRHI = VSSTNRHI
  )
```

## Derive Normal Range Flags {#anrfl}

The `derive_var_anrind()` function derives the analysis range indicator
(`ANRIND`):

```{r anrind}
advs <- advs |>
  derive_var_anrind()

advs |>
  select(USUBJID, PARAMCD, AVAL, ANRLO, ANRHI, ANRIND) |>
  head(8)
```

## Add ADSL Variables {#adsl_vars}

If needed, the remaining `ADSL` variables can now be added:

```{r adsl-vars}
advs <- advs |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTP = TRT01P, TRTA = TRT01A),
    by_vars     = exprs(STUDYID, USUBJID)
  )
```

## Derive Analysis Sequence Number {#aseq}

```{r aseq}
advs <- advs |>
  derive_var_obs_number(
    by_vars    = exprs(STUDYID, USUBJID),
    order      = exprs(PARAMCD, ADT, VISITNUM, ATPTN),
    new_var    = ASEQ,
    check_type = "error"
  )

advs |>
  select(USUBJID, PARAMCD, ADT, VISITNUM, ASEQ) |>
  head(5)
```

## Add Labels and Attributes {#attributes}

See the equivalent section in [Creating ADSL](adsl.qmd#attributes) for a
discussion of packages like `{metacore}`, `{metatools}`, and `{xportr}` for
applying metadata.

## Example Script

A template script for ADVS is available in the `{admiral}` package:

```r
use_ad_template("ADVS")
```

## See Also

-   [Creating ADSL](adsl.qmd)
-   [Date/Time Imputation](imputation.qmd)
-   [Visit and Period Variables](visits_periods.qmd)
-   [Hy's Law](hys_law.qmd)
-   [Lab Grading](lab_grading.qmd)
