# Creating a BDS Time-to-Event ADaM

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter describes creating a BDS (Basic Data Structure) Time-to-Event
(TTE) ADaM. Examples are presented and tested in the context of `ADTTE`
(Time-to-Event dataset), which is commonly derived for analyses such as overall
survival, time to response, or time to first adverse event.

::: callout-note
All examples assume CDISC SDTM and/or ADaM format as input unless otherwise
specified.
:::

## Programming Workflow

1.  [Read in Data](#readdata)
2.  [Derive Parameters](#params)
3.  [Derive Analysis Value (`AVAL`)](#aval)
4.  [Derive Censoring Indicator (`CNSR`)](#cnsr)
5.  [Add Treatment and Period Variables](#trt)
6.  [Add ADSL Variables](#adsl_vars)
7.  [Derive Analysis Sequence Number (`ASEQ`)](#aseq)
8.  [Add Labels and Attributes](#attributes)

## Read in Data {#readdata}

For example purposes, the CDISC Pilot SDTM datasets — which are included in
`{pharmaversesdtm}` — are used.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
```

```{r load-data}
ae   <- pharmaversesdtm::ae   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl
```

```{r filter-data}
ae <- ae |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:6])
```

## Derive Parameters {#params}

The function `derive_param_tte()` is used to derive TTE parameters. Each
parameter represents a different event of interest.

For `ADTTE`, typical parameters include:

-   `OS` — Overall Survival
-   `PFS` — Progression-Free Survival
-   `TTAE` — Time to First Adverse Event

```{r params}
# Define the event source (first adverse event)
ae_start_date <- event_source(
  dataset_name = "ae",
  date         = AESTDTC,
  set_values_to = exprs(
    EVNTDESC = AETERM,
    SRCDOM   = "AE",
    SRCVAR   = "AESTDTC",
    SRCSEQ   = AESEQ
  )
)

# Define the censoring source (end of study or death)
adsl_censor <- censor_source(
  dataset_name  = "adsl",
  date          = EOSDT,
  set_values_to = exprs(
    EVNTDESC = "END OF STUDY",
    SRCDOM   = "ADSL",
    SRCVAR   = "EOSDT"
  )
)

# Derive the TTE parameter
adtte <- derive_param_tte(
  dataset_adsl    = adsl,
  source_datasets = list(adsl = adsl, ae = ae),
  start_date      = TRTSDT,
  event_conditions = list(ae_start_date),
  censor_conditions = list(adsl_censor),
  set_values_to   = exprs(
    PARAMCD = "TTAE",
    PARAM   = "Time to First Adverse Event"
  )
)

adtte |>
  select(USUBJID, PARAMCD, PARAM, ADT, STARTDT, EVNTDESC, SRCDOM, SRCVAR) |>
  head(5)
```

## Derive Analysis Value {#aval}

The analysis value `AVAL` for TTE analyses represents time in the required
units (days, weeks, months, etc.):

```{r aval}
adtte <- adtte |>
  derive_vars_duration(
    new_var    = AVAL,
    start_date = STARTDT,
    end_date   = ADT
  )

adtte |>
  select(USUBJID, PARAMCD, STARTDT, ADT, AVAL) |>
  head(5)
```

## Derive Censoring Indicator {#cnsr}

The censoring indicator `CNSR` takes a value of `0` for events and `1` for
censored observations. It is derived automatically by `derive_param_tte()`.

```{r cnsr}
adtte |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC) |>
  head(5)
```

## Add Treatment and Period Variables {#trt}

Treatment variables are typically merged from `ADSL`:

```{r trt}
adtte <- adtte |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTP = TRT01P, TRTA = TRT01A),
    by_vars     = exprs(STUDYID, USUBJID)
  )
```

## Add ADSL Variables {#adsl_vars}

```{r adsl-vars}
adtte <- adtte |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(AGE, SEX, RACE, SAFFL),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adtte |>
  select(USUBJID, PARAMCD, AVAL, CNSR, TRTA, AGE, SEX, SAFFL) |>
  head(5)
```

## Derive Analysis Sequence Number {#aseq}

```{r aseq}
adtte <- adtte |>
  derive_var_obs_number(
    by_vars    = exprs(STUDYID, USUBJID),
    order      = exprs(PARAMCD),
    new_var    = ASEQ,
    check_type = "error"
  )

adtte |>
  select(USUBJID, PARAMCD, ASEQ) |>
  head(5)
```

## Add Labels and Attributes {#attributes}

See the equivalent section in [Creating ADSL](adsl.qmd#attributes) for a
discussion of packages like `{metacore}`, `{metatools}`, and `{xportr}` for
applying metadata.

## Example Script

An ADTTE template script is available in the `{admiral}` package:

```r
use_ad_template("ADTTE")
```

## See Also

-   [Creating ADSL](adsl.qmd)
-   [Date/Time Imputation](imputation.qmd)
-   [Estimands](estimands.qmd)
