# Creating a BDS Exposure ADaM

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter describes creating a BDS (Basic Data Structure) Exposure ADaM.
Examples are presented and tested in the context of `ADEX` (Exposure dataset),
which is based on the `EX` SDTM domain.

::: callout-note
All examples assume CDISC SDTM and/or ADaM format as input unless otherwise
specified.
:::

## Programming Workflow

1.  [Read in Data](#readdata)
2.  [Derive/Impute Start and End Date/Time of Exposure](#datetime)
3.  [Assign `PARAMCD`, `PARAM`, `PTYPE`](#paramcd)
4.  [Derive Analysis Value (`AVAL`)](#aval)
5.  [Derive Timing Variables](#timing)
6.  [Add Treatment Variables](#trt)
7.  [Add ADSL Variables](#adsl_vars)
8.  [Derive Analysis Sequence Number (`ASEQ`)](#aseq)
9.  [Add Labels and Attributes](#attributes)

## Read in Data {#readdata}

For example purposes, the CDISC Pilot SDTM datasets — which are included in
`{pharmaversesdtm}` — are used.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
```

```{r load-data}
ex   <- pharmaversesdtm::ex   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl
```

```{r filter-data}
ex <- ex |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:6])
```

## Derive/Impute Start and End Date/Time of Exposure {#datetime}

The function `derive_vars_dtm()` can be used to derive `EXSTDTM` and `EXENDTM`.

```{r datetime}
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST"
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    time_imputation = "last"
  )

ex_ext |>
  select(USUBJID, EXTRT, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  head(5)
```

## Assign `PARAMCD`, `PARAM`, `PTYPE` {#paramcd}

Exposure parameters can be derived from the `EX` data.
Typical exposure parameters include:

-   `DOSE` — planned dose amount
-   `DURD` — duration of exposure
-   `TNDOSE` — total number of doses

A parameter look-up table can be used to map them:

```{r paramcd}
param_lookup <- tribble(
  ~PTYPE,     ~PARAMCD, ~PARAM,
  "PLANNED",  "DOSEP",  "Planned Dose Amount (mg)",
  "ACTUAL",   "DOSEA",  "Actual Dose Amount (mg)"
)

adex <- bind_rows(
  ex_ext |> mutate(PTYPE = "PLANNED", AVAL = EXDOSE),
  ex_ext |> mutate(PTYPE = "ACTUAL",  AVAL = EXDOSE)
) |>
  derive_vars_merged(
    dataset_add = param_lookup,
    by_vars     = exprs(PTYPE)
  )

adex |>
  select(USUBJID, EXTRT, PTYPE, PARAMCD, PARAM, AVAL) |>
  head(5)
```

## Derive Analysis Value {#aval}

For exposure ADaMs, `AVAL` is typically the dose amount, number of doses, or
duration. For example, to derive a cumulative dose parameter:

```{r aval}
# Derive duration in days
adex <- adex |>
  derive_vars_duration(
    new_var      = EXDURD,
    new_var_unit = EXDURDU,
    start_date   = EXSTDTM,
    end_date     = EXENDTM,
    in_unit      = "days",
    out_unit     = "days",
    add_one      = TRUE
  )

adex |>
  select(USUBJID, EXTRT, EXSTDTM, EXENDTM, EXDURD) |>
  head(5)
```

## Derive Timing Variables {#timing}

```{r timing}
adex <- adex |>
  derive_vars_dtm_to_dt(exprs(EXSTDTM, EXENDTM)) |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(EXSTDT)
  )

adex |>
  select(USUBJID, EXSTDTM, EXSTDT, EXSTDY) |>
  head(5)
```

## Add Treatment Variables {#trt}

```{r trt}
adex <- adex |>
  mutate(TRTP = EXTRT, TRTA = EXTRT)
```

## Add ADSL Variables {#adsl_vars}

```{r adsl-vars}
adex <- adex |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(AGE, SEX, RACE),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adex |>
  select(USUBJID, PARAMCD, AVAL, AGE, SEX, RACE) |>
  head(5)
```

## Derive Analysis Sequence Number {#aseq}

```{r aseq}
adex <- adex |>
  derive_var_obs_number(
    by_vars    = exprs(STUDYID, USUBJID),
    order      = exprs(PARAMCD, EXSTDTM, EXSEQ),
    new_var    = ASEQ,
    check_type = "error"
  )

adex |>
  select(USUBJID, PARAMCD, EXSTDTM, EXSEQ, ASEQ) |>
  head(5)
```

## Add Labels and Attributes {#attributes}

See the equivalent section in [Creating ADSL](adsl.qmd#attributes) for a
discussion of packages like `{metacore}`, `{metatools}`, and `{xportr}` for
applying metadata.

## Example Script

An ADEX template script is available in the `{admiral}` package:

```r
use_ad_template("ADEX")
```

## See Also

-   [Creating ADSL](adsl.qmd)
-   [Date/Time Imputation](imputation.qmd)
-   [Creating a BDS Findings ADaM](bds_finding.qmd)
