# Programming Concepts and Conventions

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dplyr)
library(rlang)
library(admiral)
```

## Introduction

This chapter discusses the common programming concepts and conventions adopted
within the `{admiral}` family of packages. It is intended to be a user-facing
companion to the
[Programming Strategy](https://pharmaverse.github.io/admiraldev/articles/programming_strategy.html)
developer guide.

For common `{admiral}` frequently asked questions, see the [FAQ](faq.qmd)
chapter.

## Input and Output

It is expected that the input dataset is **not grouped**. Otherwise an error is
issued.

The output dataset is **ungrouped**. The observations are not ordered in a
dedicated way. In particular, the order of the observations of the input dataset
may not be preserved.

## `{admiral}` Functions and Options

As a general principle, the behaviour of the `{admiral}` functions is only
determined by their input, not by any global object — i.e. all inputs like
datasets, variable names, and options must be provided to the function via
arguments. Correspondingly, functions do not have any side-effects like creating
or modifying global objects, printing, or writing files.

An exception to the above principle is our approach to package options (see
`get_admiral_option()` and `set_admiral_options()`), which allow for
user-defined defaults on commonly used function arguments. For instance, the
option `subject_keys` is pre-defined as `exprs(STUDYID, USUBJID)`, but can be
modified using `set_admiral_options(subject_keys = exprs(...))` at the top of a
script.

## Handling of Missing Values {#missing}

When using the `{haven}` package to read SAS datasets into R, SAS-style
character missing values (i.e. `""`) are **not** converted into proper R `NA`
values. Rather they are kept as empty strings. This is problematic for any
downstream data processing because R handles `""` just like any other string.
Thus, before any data manipulation, SAS blanks should be converted to R `NA`s
using `{admiral}`'s `convert_blanks_to_na()` function:

```r
dm <- haven::read_sas("dm.sas7bdat") |>
  convert_blanks_to_na()
```

Note that any logical operator applied to an `NA` value **always** returns `NA`
rather than `TRUE` or `FALSE`.

```{r na-logical}
visits <- c("Baseline", NA, "Screening", "Week 1 Day 7")
visits != "Baseline"
```

The only exception is `is.na()`, which returns `TRUE` if the input is `NA`.

```{r is-na}
is.na(visits)
```

Thus, to filter all visits which are not `"Baseline"` the following condition
must be used:

```{r filter-na}
visits != "Baseline" | is.na(visits)
```

Also note that most aggregation functions, like `mean()` or `max()`, also return
`NA` if any element of the input vector is missing.

```{r mean-na}
mean(c(1, NA, 2))
```

To avoid this behaviour you must explicitly set `na.rm = TRUE`.

```{r mean-narm}
mean(c(1, NA, 2), na.rm = TRUE)
```

This is very important to keep in mind when using `{admiral}`'s aggregation
functions such as `derive_summary_records()`.

## Expressions in Scripts {#exprs}

### Quoting and Unquoting: `expr()`, `exprs()`, `!!` and `!!!`

#### `expr()` and `exprs()`

`expr()` is a function from the `{rlang}` package, which is used to create an
**expression**. The expression is not evaluated — rather, it is passed to the
derivation function, which evaluates it in its own environment. `exprs()` is
the plural version: it accepts multiple comma-separated items and returns a
list of expressions.

```{r expr-demo}
library(rlang)

adae <- data.frame(USUBJID = "XXX-1", AEDECOD = "HEADACHE")

# Return the adae object
adae

# Return an expression
expr(adae)
```

When used within an `{admiral}` derivation function, `expr()` and `exprs()`
allow the function to evaluate the expressions in the context of the input
dataset. As a result, users can pass variable names of datasets to the function
without wrapping them in quotation marks.

The expressions framework is powerful because users are able to intuitively
"inject code" into `{admiral}` functions (through the function parameters)
using very similar syntax to open code, with the exception being an outer
`exprs()` wrapper. For instance, in the `derive_vars_merged()` call below, the
user is merging `adsl` with `ex` and is able to filter `ex` prior to the merge
using an expression passed to the `filter_add` parameter:

```r
derive_vars_merged(
  adsl,
  dataset_add = ex,
  filter_add  = !is.na(EXENDTM),
  by_vars     = exprs(STUDYID, USUBJID),
  new_vars    = exprs(
    TRTEDTM  = EXENDTM,
    TRTETMF  = EXENTMF,
    COMPTRT  = if_else(!is.na(EXENDTM), "Y", "N")
  ),
  order = exprs(EXENDTM),
  mode  = "last"
)
```

#### Bang-Bang (`!!`) and Bang-Bang-Bang (`!!!`) {#unquoting}

Sometimes you may want to construct an expression using other, pre-existing
expressions. However, it's not immediately clear how to achieve this because
expressions inherently pause evaluation of your code before it is executed:

```{r bang-bang-intro}
a <- expr(2)
b <- expr(3)

expr(a + b)
# NOT 2 + 3
```

This is where `!!` (bang-bang) comes in. Provided by the `{rlang}` package, it
allows you to inject the contents of an expression into another expression.
By using `!!` you are **unquoting** an expression, i.e. evaluating it before
you pass it onwards.

```{r bang-bang}
expr(!!a + !!b)
```

`!!!` (bang-bang-bang) is the plural version of `!!` and can be used to
unquote a list of expressions:

```{r bang-bang-bang}
exprs(!!!list(a, b))
```

Within `{admiral}`, this operator can be useful when you need to unquote a
list of variables stored as expressions. One example is the `{admiral}` subject
keys:

```{r subject-keys}
get_admiral_option("subject_keys")
```

If we want to use the subject keys stored within this `{admiral}` option to
subset a dataset, we need to use `!!!` to unquote this list:

```{r subject-keys-select}
adcm <- data.frame(STUDYID = "XXX", USUBJID = "XXX-1", CMTRT = "ASPIRIN")
adcm

# This doesn't work — we are not unquoting the subject keys
try(adcm |> select(get_admiral_option("subject_keys")), silent = TRUE)

# This works because we are unquoting the subject keys
adcm |> select(!!!get_admiral_option("subject_keys"))
```

#### Summary

The expressions framework may seem slightly unfamiliar at first, but it allows
for significant power and flexibility. For a comprehensive treatment of
expressions, see
[Chapter 18](https://adv-r.hadley.nz/expressions.html) and
[Chapter 19](https://adv-r.hadley.nz/quasiquotation.html) of *Advanced R*.
Chapter 19 specifically covers the concept of unquoting in much more detail.

### Common Pitfalls {#pitfalls}

Expressions are powerful, but they can also lead to misunderstandings. Let's
set up some dummy data to explore common issues.

```{r pitfall-data}
vs <- tribble(
  ~USUBJID,  ~VSTESTCD, ~VISIT,       ~VSSTRESN, ~VSSTRESU, ~VSDTC,
  "01-1301", "WEIGHT",  "SCREENING",  82.1,      "kg",      "2013-08-29",
  "01-1301", "WEIGHT",  "WEEK 2",     81.19,     "kg",      "2013-09-15",
  "01-1301", "WEIGHT",  "WEEK 4",     82.56,     "kg",      "2013-09-24",
  "01-1302", "BMI",     "SCREENING",  20.1,      "kg/m2",   "2013-08-29",
  "01-1302", "BMI",     "WEEK 2",     20.2,      "kg/m2",   "2013-09-15",
  "01-1302", "BMI",     "WEEK 4",     19.9,      "kg/m2",   "2013-09-24"
)

dm <- tribble(
  ~USUBJID,  ~AGE,
  "01-1301", 18
)
```

#### Pitfall 1: Passing an object name instead of an expression {#pitfall1}

When writing more complex `{admiral}` code it can be easy to mistakenly pass
the wrong input to an argument that expects an expression. The code below fails
because `my_expression` is not an expression — it is the **name** of an object
in the global environment *containing* an expression.

```{r pitfall1-fail}
#| error: true
my_expression <- expr(VSTESTCD == "WEIGHT" & VISIT == "SCREENING")

derive_vars_merged(
  dm,
  dataset_add = select(vs, USUBJID, VSTESTCD, VISIT),
  by_vars     = exprs(USUBJID),
  filter_add  = my_expression
)
```

To fix this code, we need to [unquote](#unquoting) `my_expression` so that the
expression it holds is passed correctly:

```{r pitfall1-fix}
derive_vars_merged(
  dm,
  dataset_add = select(vs, USUBJID, VSTESTCD, VISIT),
  by_vars     = exprs(USUBJID),
  filter_add  = !!my_expression
)
```

#### Pitfall 2: Expressions referencing variables that don't exist

Even if an actual expression *is* passed, you must make sure it can be
evaluated within the dataset of interest. This may seem trivial, but it is a
common pitfall because expressions delay evaluation of code and can therefore
delay the identification of issues:

```{r pitfall2}
#| error: true
filter_vs_and_merge <- function(my_expression) {
  derive_vars_merged(
    dm,
    dataset_add = select(vs, USUBJID, VSTESTCD, VISIT),
    by_vars     = exprs(USUBJID),
    filter_add  = !!my_expression
  )
}

# This works
filter_vs_and_merge(expr(VSTESTCD == "WEIGHT" & VISIT == "SCREENING"))

# This fails because VSTPT was dropped from vs inside filter_vs_and_merge()
filter_vs_and_merge(
  expr(VSTESTCD == "WEIGHT" & VISIT == "SCREENING" & VSTPT == "PREDOSE")
)
```

## See Also

-   [Programming Strategy](https://pharmaverse.github.io/admiraldev/articles/programming_strategy.html)
-   [FAQ](faq.qmd)
