# Estimands

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

The ICH E9(R1) addendum introduced the concept of **estimands** to align
the clinical question of interest with the analysis. The estimand
framework requires that intercurrent events (ICEs) — events occurring
after treatment initiation that can affect the interpretation of the
outcome — are pre-specified and handled according to one of the defined
strategies (treatment policy, composite, while-on-treatment, hypothetical,
or principal stratum).

In ADaM terms, these strategies are implemented by creating ICE flag
variables that can be used to subset or weight analyses. The `{admiral}`
package provides functions to derive these flag variables.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
library(tibble)
```

## Common Intercurrent Event Flags

### Treatment Policy Strategy

Under the treatment policy strategy, observations are included regardless
of whether an ICE occurred. No additional derivation is required —
all observations are used.

### While-on-Treatment Strategy

The while-on-treatment strategy includes only observations collected while
the subject was on treatment. The flag `ONTRTFL` can be used:

```{r ontrtfl}
adsl <- admiral::admiral_adsl
vs   <- pharmaversesdtm::vs |> convert_blanks_to_na()

advs <- vs |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:4]) |>
  derive_vars_dt(new_vars_prefix = "A", dtc = VSDTC) |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT, TRTEDT),
    by_vars     = exprs(STUDYID, USUBJID)
  )

advs <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT
  )

advs |>
  select(USUBJID, VSTESTCD, ADT, TRTSDT, TRTEDT, ONTRTFL) |>
  head(8)
```

### Composite Strategy

The composite strategy uses an alternative value (e.g. worst outcome or
`NA`) for observations after an ICE. The ICE date is typically merged
from `ADSL` or derived from `DS`.

```{r composite}
# Example: flag post-discontinuation records
ds <- pharmaversesdtm::ds |> convert_blanks_to_na()

# Get discontinuation dates
disc_dates <- ds |>
  filter(DSDECOD == "STUDY DISCONTINUATION") |>
  derive_vars_dt(new_vars_prefix = "DISC", dtc = DSSTDTC) |>
  select(STUDYID, USUBJID, DISCDT)

advs <- advs |>
  derive_vars_merged(
    dataset_add = disc_dates,
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(
    POSTDISCFL = if_else(!is.na(DISCDT) & ADT > DISCDT, "Y", NA_character_)
  )

advs |>
  filter(!is.na(POSTDISCFL)) |>
  select(USUBJID, VSTESTCD, ADT, DISCDT, POSTDISCFL) |>
  head(5)
```

### Hypothetical Strategy

The hypothetical strategy considers outcomes as if the ICE had not
occurred. This is typically implemented through sensitivity analyses
(e.g. multiple imputation). The flag variable identifies records where
the ICE occurred so that they can be treated appropriately in the model.

```{r hypothetical}
advs <- advs |>
  mutate(
    ICEFLAG = if_else(!is.na(DISCDT) & ADT >= DISCDT, "Y", NA_character_)
  )

advs |>
  filter(!is.na(ICEFLAG)) |>
  select(USUBJID, VSTESTCD, ADT, DISCDT, ICEFLAG) |>
  head(5)
```

## Deriving Rescue Medication Use Flag

A common ICE in many trials is the use of rescue medication. The flag
can be derived using `derive_var_merged_exist_flag()`:

```{r rescue-med}
cm <- tibble::tribble(
  ~STUDYID, ~USUBJID, ~CMTRT,       ~CMSTDTC,
  "A",      "1",      "RESCUE MED", "2022-03-15",
  "A",      "2",      "IBUPROFEN",  "2022-04-01"
)

adsl_ex <- tibble::tribble(
  ~STUDYID, ~USUBJID,
  "A",      "1",
  "A",      "2",
  "A",      "3"
)

adsl_ex <- adsl_ex |>
  derive_var_merged_exist_flag(
    dataset_add = cm,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = RESCUEFL,
    condition   = CMTRT == "RESCUE MED"
  )

adsl_ex
```

## Notes on Implementation

-   The choice of ICE flag names (e.g. `ONTRTFL`, `POSTDISCFL`,
    `ICEFLAG`) should follow the study SAP and your organisation's ADaM
    programming conventions.
-   Multiple ICEs may need to be handled within the same dataset; each
    should have a dedicated flag variable.
-   Consult the ICH E9(R1) addendum for the full definition of each
    estimand strategy.

## See Also

-   [Creating ADSL](adsl.qmd)
-   [Creating a BDS Time-to-Event ADaM](bds_tte.qmd)
-   [Creating a BDS Findings ADaM](bds_finding.qmd)
