# Frequently Asked Questions (FAQ)

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(admiral)
library(dplyr)
```

## General Questions

### What is `{admiral}`?

`{admiral}` (**A**DaM in **R** **A**sset **L**ibrary) is an open-source R
package for the creation of ADaM datasets. It provides a modular collection
of functions that follow CDISC ADaM standards. The package is developed and
maintained by the [pharmaverse](https://pharmaverse.org/) community.

### Where can I get help?

-   [admiral Slack channel](https://pharmaverse.slack.com/) — post in
    `#admiral` for questions and discussion.
-   [GitHub Issues](https://github.com/pharmaverse/admiral/issues) — to
    report bugs or request new features.
-   [This book](https://pharmaverse.github.io/admiral-book/) — for
    worked examples and concepts.

### How do I get started?

Start with the [Introduction](introduction.qmd) chapter of this book,
which walks through the main ideas of `{admiral}` with worked examples.
Then work through the [ADaM Data Structures](adam_structures.qmd) chapters
for the dataset types you need.

You can also use the template scripts provided by `{admiral}`:

```{r list-templates}
list_all_templates()
```

---

## Programming Questions

### Why do `{admiral}` functions use `exprs()` instead of character strings?

`{admiral}` uses the `{rlang}` expressions framework so that variable
names can be used directly (without quotes) in function arguments. This
makes code more readable and allows `{admiral}` functions to evaluate
expressions in the context of the input dataset.

For a detailed explanation, see the
[Expressions in Scripts](concepts_conventions.qmd#exprs) section of the
[Programming Concepts and Conventions](concepts_conventions.qmd) chapter.

### How do I handle missing values from SAS datasets?

When reading SAS datasets with `{haven}`, SAS character missing values
(`""`) are **not** automatically converted to R `NA`. You must call
`convert_blanks_to_na()` on each dataset before any `{admiral}` processing:

```r
dm <- haven::read_sas("dm.sas7bdat") |>
  convert_blanks_to_na()
```

See [Handling of Missing Values](concepts_conventions.qmd#missing) for
full details.

### Why does `filter_add = my_expression` fail?

If `my_expression` is an R object *containing* an expression (rather than
an expression itself), you need to unquote it with `!!`:

```{r faq-expressions}
#| error: true
library(rlang)
my_expression <- expr(VSTESTCD == "WEIGHT")

vs <- dplyr::tribble(
  ~USUBJID, ~VSTESTCD, ~VISIT,
  "01",     "WEIGHT",  "SCREENING"
)
dm <- dplyr::tribble(~USUBJID, ~AGE, "01", 18)

# This fails:
try(
  derive_vars_merged(
    dm,
    dataset_add = select(vs, USUBJID, VSTESTCD, VISIT),
    by_vars     = exprs(USUBJID),
    filter_add  = my_expression
  ),
  silent = TRUE
)

# This works:
derive_vars_merged(
  dm,
  dataset_add = select(vs, USUBJID, VSTESTCD, VISIT),
  by_vars     = exprs(USUBJID),
  filter_add  = !!my_expression
)
```

For more details, see [Common Pitfalls](concepts_conventions.qmd#pitfalls).

### How do I apply a derivation to only some records?

Use `restrict_derivation()` to apply a derivation only to records matching
a filter condition:

```r
adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(USUBJID),
      order   = exprs(ASTDTM, AESEQ),
      new_var = AOCCIFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )
```

See [Higher Order Functions](higher_order.qmd) for details.

### How do I apply different imputation rules for different records?

Use `slice_derivation()` to apply different arguments to different
subsets of the data:

```r
vs |>
  slice_derivation(
    derivation = derive_vars_dtm,
    args = params(dtc = VSDTC, new_vars_prefix = "A"),
    derivation_slice(
      filter = VSTPT == "PRE-DOSE",
      args   = params(time_imputation = "first")
    ),
    derivation_slice(
      filter = TRUE,
      args   = params(time_imputation = "last")
    )
  )
```

See [Date/Time Imputation](imputation.qmd) and
[Higher Order Functions](higher_order.qmd) for details.

---

## Data and Package Questions

### What test data does `{admiral}` use?

`{admiral}` uses SDTM and ADaM datasets from the
[`{pharmaversesdtm}`](https://pharmaverse.github.io/pharmaversesdtm/)
package, which contains data from the CDISC Pilot Study. These datasets
are used in all worked examples in this book.

### Does `{admiral}` support CDISC standards?

Yes. `{admiral}` is designed to follow the CDISC ADaM Implementation
Guide (ADaMIG v1.3). Each derivation function is designed to produce
output consistent with the ADaMIG requirements.

### How do I add metadata and export to XPT?

Use the following packages as a final step after `{admiral}` derivations:

-   [`{metacore}`](https://atorus-research.github.io/metacore/) — manage
    metadata specifications.
-   [`{metatools}`](https://pharmaverse.github.io/metatools/) — apply
    metadata to datasets.
-   [`{xportr}`](https://atorus-research.github.io/xportr/) — convert
    R data frames to transport v5 (XPT) files.

See the [pharmaverse E2E example](https://pharmaverse.github.io/examples/)
for a complete end-to-end workflow.

---

## See Also

-   [Introduction to admiral](introduction.qmd)
-   [Programming Concepts and Conventions](concepts_conventions.qmd)
-   [Higher Order Functions](higher_order.qmd)
-   [pharmaverse website](https://pharmaverse.org/)
