# Generic Derivations

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`{admiral}` provides a set of general-purpose derivation functions that
can be applied across many different ADaM datasets and derivation
scenarios. This chapter covers the most commonly used generic functions.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(tibble)
library(lubridate)
library(pharmaversesdtm)
```

## `derive_vars_merged()` — Merge Variables from Another Dataset

`derive_vars_merged()` adds variable(s) to the input dataset from a
look-up or reference dataset based on a by-variable key.

```{r derive-vars-merged}
adsl <- admiral::admiral_adsl
ae   <- pharmaversesdtm::ae |> convert_blanks_to_na()

adae <- ae |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:4]) |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT, TRTEDT, TRT01A, TRT01P),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adae |>
  select(USUBJID, AETERM, TRTSDT, TRTEDT, TRT01A) |>
  head(5)
```

## `derive_vars_joined()` — Merge Based on a Condition

`derive_vars_joined()` is similar to `derive_vars_merged()` but also
allows filtering based on conditions that reference both datasets. It is
used when the records to merge depend on date or value comparisons.

```{r derive-vars-joined}
# Derive the last dose date before the start of each AE
ex    <- pharmaversesdtm::ex |> convert_blanks_to_na()
ex_dt <- ex |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:4]) |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST", flag_imputation = "none")

adae <- adae |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC, highest_imputation = "M") |>
  derive_vars_joined(
    ex_dt,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(LDOSEDTM = EXSTDTM),
    join_vars   = exprs(EXSTDTM),
    join_type   = "all",
    order       = exprs(EXSTDTM),
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & grepl("PLACEBO", EXTRT))) & !is.na(EXSTDTM),
    filter_join = EXSTDTM <= ASTDTM,
    mode        = "last"
  )

adae |>
  select(USUBJID, AETERM, AESTDTC, ASTDTM, LDOSEDTM) |>
  head(5)
```

## `derive_summary_records()` — Add Summary Records

`derive_summary_records()` adds new records to the dataset based on
aggregations across by-group records. It is commonly used to derive
summary parameters (e.g. average, maximum) in BDS datasets.

```{r derive-summary-records}
vs <- pharmaversesdtm::vs |> convert_blanks_to_na()

advs <- vs |>
  filter(
    USUBJID %in% unique(adsl$USUBJID)[1:4],
    VSTESTCD %in% c("SYSBP", "DIABP")
  ) |>
  mutate(PARAMCD = VSTESTCD, AVAL = VSSTRESN) |>
  derive_vars_dt(new_vars_prefix = "A", dtc = VSDTC)

# Add a mean AVAL summary record per subject and visit
advs_with_mean <- advs |>
  derive_summary_records(
    by_vars       = exprs(STUDYID, USUBJID, VISIT, VISITNUM),
    filter_add    = PARAMCD %in% c("SYSBP", "DIABP"),
    set_values_to = exprs(
      AVAL    = mean(AVAL, na.rm = TRUE),
      PARAMCD = "MABP",
      PARAM   = "Mean Arterial Blood Pressure"
    )
  )

advs_with_mean |>
  filter(PARAMCD == "MABP") |>
  select(USUBJID, VISIT, PARAMCD, PARAM, AVAL) |>
  head(5)
```

## `derive_var_obs_number()` — Observation Number

`derive_var_obs_number()` assigns a sequential observation number
(`ASEQ` or similar) within a by group:

```{r aseq}
advs |>
  derive_var_obs_number(
    by_vars    = exprs(STUDYID, USUBJID),
    order      = exprs(PARAMCD, ADT, VISITNUM),
    new_var    = ASEQ,
    check_type = "error"
  ) |>
  select(USUBJID, PARAMCD, ADT, VISITNUM, ASEQ) |>
  head(8)
```

## `derive_var_extreme_flag()` — Extreme Record Flag

`derive_var_extreme_flag()` flags the first or last record within a
by-group according to a sort order. Useful for flagging the last record
before treatment start (baseline) or the first occurrence of an event.

```{r extreme-flag}
advs |>
  filter(!is.na(AVAL)) |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, USUBJID, PARAMCD),
    order   = exprs(ADT, VISITNUM),
    new_var = LASTFL,
    mode    = "last"
  ) |>
  select(USUBJID, PARAMCD, ADT, AVAL, LASTFL) |>
  head(8)
```

## Sort Order {#sort_order}

When `NA` values appear in sorting variables used by `{admiral}`
functions (e.g. the `order` argument of `derive_var_extreme_flag()`),
they are treated as the **smallest** value by default. Use `desc()` to
sort in descending order; `NA` values will then be treated as the
**largest**.

```{r sort-order-na}
# Illustrate NA sort order
advs |>
  filter(PARAMCD == "SYSBP") |>
  derive_var_extreme_flag(
    by_vars = exprs(USUBJID),
    order   = exprs(ADT),   # NAs go first
    new_var = FIRSTFL,
    mode    = "first"
  ) |>
  select(USUBJID, ADT, AVAL, FIRSTFL) |>
  head(6)
```

## See Also

-   [Introduction to admiral](introduction.qmd)
-   [Higher Order Functions](higher_order.qmd)
-   [Date/Time Imputation](imputation.qmd)
