# Higher Order Functions

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Higher order functions are `{admiral}` functions that take other
functions as input. They enhance the existing portfolio of derivation
functions by allowing greater customisation of the latter's behaviour.
A derivation function can be:

-   `call_derivation()` — Called multiple times, while varying some of
    the input arguments.
-   `restrict_derivation()` — Executed on a subset of the input dataset.
-   `slice_derivation()` — Executed differently on subsets of the input
    dataset.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(tibble)
```

## `call_derivation()` — Call a Function Multiple Times

`call_derivation()` is useful when the same derivation must be applied
multiple times with slightly different arguments. For example, when
deriving multiple BDS findings parameters that share the same base
derivation logic:

```{r call-derivation}
vs   <- pharmaversesdtm::vs   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl

advs <- vs |>
  filter(
    USUBJID %in% unique(adsl$USUBJID)[1:4],
    VSTESTCD %in% c("SYSBP", "DIABP", "PULSE", "WEIGHT")
  ) |>
  mutate(
    PARAMCD = VSTESTCD,
    PARAM   = VSTEST,
    AVAL    = VSSTRESN
  ) |>
  derive_vars_dt(new_vars_prefix = "A", dtc = VSDTC) |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT),
    by_vars     = exprs(STUDYID, USUBJID)
  )

# Derive ABLFL and BASE for multiple parameters using call_derivation()
advs <- call_derivation(
  derivation = restrict_derivation,
  args = params(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )
)(advs)

advs <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  )

advs |>
  select(USUBJID, PARAMCD, ADT, AVAL, ABLFL, BASE) |>
  filter(!is.na(BASE)) |>
  head(8)
```

## `restrict_derivation()` — Apply to a Subset

`restrict_derivation()` allows a derivation to be executed only on a
subset of records (defined by `filter`). Records not matching the filter
are kept but unchanged.

```{r restrict-derivation}
# Flag the first occurrence of most severe AE, restricted to TEAEs only
ae <- pharmaversesdtm::ae |> convert_blanks_to_na()

adae <- ae |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:4]) |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT, TRTEDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC) |>
  mutate(ASEV = AESEV) |>
  derive_var_trtemfl(
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 30
  )

adae <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(
        as.integer(factor(
          ASEV,
          levels = c("DEATH THREATENING", "SEVERE", "MODERATE", "MILD")
        )),
        ASTDT, AESEQ
      ),
      new_var = AOCCIFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

adae |>
  select(USUBJID, AESEQ, ASEV, ASTDT, TRTEMFL, AOCCIFL) |>
  head(8)
```

## `slice_derivation()` — Apply Different Logic to Subsets

`slice_derivation()` splits the input dataset into slices, applies a
potentially different derivation to each slice, and combines the
results. This is useful when different records need different imputation
or derivation logic.

```{r slice-derivation}
# Use different time imputation for pre-dose vs other VS records
vs_timed <- vs |>
  filter(
    USUBJID %in% unique(adsl$USUBJID)[1:2],
    VSTESTCD == "SYSBP"
  ) |>
  slice_derivation(
    derivation = derive_vars_dtm,
    args = params(
      dtc             = VSDTC,
      new_vars_prefix = "A"
    ),
    derivation_slice(
      filter = !is.na(VSTPT) & toupper(VSTPT) == "PRE-DOSE",
      args   = params(time_imputation = "first")
    ),
    derivation_slice(
      filter = TRUE,
      args   = params(time_imputation = "last")
    )
  )

vs_timed |>
  select(USUBJID, VSTESTCD, VSDTC, VSTPT, ADTM) |>
  head(6)
```

## Summary

| Function | Use case |
|---|---|
| `call_derivation()` | Call the same function multiple times with varying arguments |
| `restrict_derivation()` | Apply a derivation to a filtered subset of records |
| `slice_derivation()` | Apply different derivations to different subsets of records |

## See Also

-   [Introduction to admiral](introduction.qmd)
-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [Date/Time Imputation](imputation.qmd)
