# Hy's Law

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Hy's Law is a clinical rule used to identify drug-induced liver injury
(DILI). The criteria require:

1.  ALT or AST ≥ 3× Upper Limit of Normal (ULN), **and**
2.  Total bilirubin ≥ 2× ULN, **and**
3.  No other explanation (e.g. hepatitis, biliary obstruction) for the
    laboratory abnormalities.

In ADaM terms, Hy's Law criteria are derived by flagging subjects and/or
records in the laboratory dataset (typically `ADLB`).

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
```

## Programming Workflow

The derivation involves three main steps:

1.  Identify records where ALT or AST ≥ 3× ULN.
2.  Identify records where Total Bilirubin ≥ 2× ULN.
3.  Identify subjects satisfying both criteria simultaneously (or within
    a study-defined time window).

### Step 1: Prepare the Lab Dataset

```{r prepare-adlb}
lb   <- pharmaversesdtm::lb   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl

adlb <- lb |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:6]) |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(
    PARAMCD = LBTESTCD,
    PARAM   = LBTEST,
    AVAL    = LBSTRESN,
    AVALU   = LBSTRESU,
    ANRHI   = LBSTNRHI
  ) |>
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc             = LBDTC
  ) |>
  filter(PARAMCD %in% c("ALT", "AST", "BILI"))

adlb |>
  select(USUBJID, PARAMCD, AVAL, ANRHI, ADT) |>
  head(10)
```

### Step 2: Derive Hy's Law Numerator Flag

Identify records where ALT or AST ≥ 3× ULN:

```{r hyslaw-alt-ast}
adlb <- adlb |>
  mutate(
    # Ratio to ULN
    AVALN_ULN = if_else(!is.na(ANRHI) & ANRHI > 0, AVAL / ANRHI, NA_real_),
    # Hy's Law component 1 flag
    HYS_NUM = if_else(
      PARAMCD %in% c("ALT", "AST") & !is.na(AVALN_ULN) & AVALN_ULN >= 3,
      "Y",
      NA_character_
    )
  )

adlb |>
  filter(!is.na(HYS_NUM)) |>
  select(USUBJID, PARAMCD, AVAL, ANRHI, AVALN_ULN, HYS_NUM) |>
  head(5)
```

### Step 3: Derive Hy's Law Denominator Flag

Identify records where Total Bilirubin ≥ 2× ULN:

```{r hyslaw-bili}
adlb <- adlb |>
  mutate(
    HYS_DEN = if_else(
      PARAMCD == "BILI" & !is.na(AVALN_ULN) & AVALN_ULN >= 2,
      "Y",
      NA_character_
    )
  )

adlb |>
  filter(!is.na(HYS_DEN)) |>
  select(USUBJID, PARAMCD, AVAL, ANRHI, AVALN_ULN, HYS_DEN) |>
  head(5)
```

### Step 4: Derive Subject-Level Hy's Law Flag

A subject satisfies Hy's Law if they have both a numerator and a
denominator flag. Use `derive_var_merged_exist_flag()` to create a
subject-level flag:

```{r hyslaw-subj}
# Subjects with ALT/AST >= 3xULN
subj_hys_num <- adlb |>
  filter(!is.na(HYS_NUM)) |>
  distinct(STUDYID, USUBJID)

# Subjects with BILI >= 2xULN
subj_hys_den <- adlb |>
  filter(!is.na(HYS_DEN)) |>
  distinct(STUDYID, USUBJID)

# Subjects satisfying both criteria (Hy's Law potential cases)
hys_subjects <- inner_join(subj_hys_num, subj_hys_den, by = c("STUDYID", "USUBJID")) |>
  mutate(HYSC_FL = "Y")

hys_subjects
```

## Notes on Study-Specific Implementation

In practice, the Hy's Law derivation must be tailored to the study:

-   The time window between the ALT/AST ≥ 3× ULN and the
    bilirubin ≥ 2× ULN events may be restricted (e.g. within the
    same study phase or within X days).
-   Baseline records should typically be excluded.
-   Results should be added to `ADLB` and/or `ADSL` as flags.

::: callout-tip
The `derive_vars_joined()` function is useful for linking ALT/AST flags
to bilirubin records that fall within a specified time window.
:::

## Example Script

An ADLB template that includes Hy's Law derivations is available:

```r
use_ad_template("ADLB")
```

## See Also

-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [Lab Grading](lab_grading.qmd)
-   [Date/Time Imputation](imputation.qmd)
