# Date and Time Imputation

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter is broken into three major sections. The first section
briefly explores the imputation rules used in `{admiral}`. The second
section focuses on imputation functions that work on vectors, with small
examples to explore the imputation rules. These **vector-based**
functions form the backbone of `{admiral}`'s more powerful functions
`derive_vars_dt()` and `derive_vars_dtm()` for building ADaM datasets.
The final section provides more detailed examples that a user might face
while working on ADaMs needing `*DT` and `*DTM` variables.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(lubridate)
library(tibble)
library(dplyr)
```

## Imputation Rules

Date and time is collected in SDTM as character values using the extended
[ISO 8601](https://en.wikipedia.org/wiki/ISO_8601) format — for example,
`"2019-10-09T13:42:00"`. The format allows some parts of the date or time
to be missing, e.g. `"2019-10"` if the day and time are unknown.

ADaM timing variables like `ADTM` or `ADY` are numeric and can only be
derived if the date or datetime is complete. Therefore `{admiral}`
provides imputation functions which fill in missing date or time parts
according to configurable imputation rules.

The key concept is **Highest Imputation**, controlled by the
`highest_imputation` argument. This argument controls which components of
the `--DTC` value are imputed if they are missing.

-   Default for `_dtm()` functions: `"h"` (hours).
-   Default for `_dt()` functions: `"n"` (no imputation).

::: callout-warning
If a component at a higher level than `highest_imputation` is missing,
`NA_character_` is returned. For example, for `highest_imputation = "D"`
a value of `"2020"` results in `NA_character_` because the month is
missing.
:::

## Vector-Based Imputation

### Basic Imputation

```{r basic-imputation}
impute_dtc_dtm(
  "2019-10",
  highest_imputation = "M",
  date_imputation    = "01-01",
  time_imputation    = "00:00:00"
)
```

Using `"first"` or `"last"` avoids invalid dates (e.g. February 31st):

```{r first-last}
impute_dtc_dtm(
  "2019-02",
  highest_imputation = "M",
  date_imputation    = "last",
  time_imputation    = "00:00:00"
)
```

The `"mid"` keyword imputes missing day to `15`, or missing day and month
to `06-30`:

```{r mid}
dates <- c("2019-02", "2019", "2019---01")

impute_dtc_dtm(
  dates,
  highest_imputation = "M",
  date_imputation    = "mid",
  time_imputation    = "00:00:00",
  preserve           = FALSE
)
```

### Imputation Level

The `highest_imputation` argument controls which components are imputed:

```{r imputation-level}
dates <- c(
  "2019-02-03T12:30:15",
  "2019-02-03T12:30",
  "2019-02-03",
  "2019-02",
  "2019"
)

# Do not impute
impute_dtc_dtm(dates, highest_imputation = "n")

# Impute time only (hours, minutes, seconds)
impute_dtc_dtm(dates, highest_imputation = "h")

# Impute days and time
impute_dtc_dtm(dates, highest_imputation = "D")

# Impute date (months and days) and time
impute_dtc_dtm(dates, highest_imputation = "M")
```

### Minimum and Maximum Dates

In some scenarios the imputed date should not be before or after certain
dates (e.g. data cut-off or death date). Use the `min_dates` and
`max_dates` arguments:

```{r min-max-dates}
impute_dtc_dtm(
  "2019-02",
  highest_imputation = "M",
  date_imputation    = "last",
  time_imputation    = "last",
  max_dates          = list(ymd("2019-01-14"), ymd("2019-02-25"))
)
```

Completely missing dates can also be imputed when `min_dates` or
`max_dates` is specified and `highest_imputation = "Y"`:

```{r year-imputation}
impute_dtc_dtm(
  c("2019-02", NA),
  highest_imputation = "Y",
  date_imputation    = "last",
  time_imputation    = "last",
  max_dates          = list(
    ymd("2019-01-14", NA),
    ymd("2019-02-25", "2020-01-01")
  )
)
```

## Imputation Flags

ADaM requires that imputed date or datetime variables are accompanied by
date/time imputation flag variables (`*DTF` and `*TMF`). These variables
indicate the highest level that was imputed. The `{admiral}` functions
that derive imputed variables also add the corresponding imputation flag
variables automatically.

::: callout-note
The `ignore_seconds_flag` argument can be set to `FALSE` when seconds
are expected, as per the ADaM IG: if only hours and minutes are ever
collected and seconds are imputed as `00`, it is not necessary to set
`*TMF` to `"S"`.
:::

## `{admiral}` Imputation Functions

| Function | Description |
|---|---|
| `derive_vars_dt()` | Derives a date variable and optional date imputation flag |
| `derive_vars_dtm()` | Derives a datetime variable plus date and time imputation flags |
| `impute_dtc_dtm()` | Returns a complete ISO 8601 datetime or `NA` |
| `impute_dtc_dt()` | Returns a complete ISO 8601 date or `NA` |
| `convert_dtc_to_dt()` | Returns a date if the ISO 8601 date is complete |
| `convert_dtc_to_dtm()` | Returns a datetime if the ISO 8601 date is complete |
| `compute_dtf()` | Returns the date imputation flag |
| `compute_tmf()` | Returns the time imputation flag |

## Examples

### Create an Imputed Datetime Variable

```{r derive-dtm}
ae <- tribble(
  ~AESTDTC,
  "2019-08-09T12:34",
  "2019-04-12",
  "2010-09",
  NA_character_
) |>
  derive_vars_dtm(
    dtc                = AESTDTC,
    new_vars_prefix    = "AST",
    highest_imputation = "M",
    date_imputation    = "first",
    time_imputation    = "first"
  ) |>
  derive_vars_dtm_to_dt(exprs(ASTDTM))

ae
```

### Create an Imputed Date Variable

```{r derive-dt}
ae2 <- tribble(
  ~AESTDTC,
  "2019-08-09T12:34:00",
  "2019-04-12",
  "2010-09",
  NA_character_
) |>
  derive_vars_dt(
    dtc                = AESTDTC,
    new_vars_prefix    = "AST",
    highest_imputation = "M",
    date_imputation    = "first"
  )

ae2
```

### Avoid Imputed Dates Before Treatment Start

To avoid imputed AE start dates before treatment start, specify
`TRTSDTM` for `min_dates`:

```{r min-dates}
ae3 <- tribble(
  ~AESTDTC,           ~TRTSDTM,
  "2019-08-09T12:34", ymd_hm("2019-11-11T12:34"),
  "2019-10",          ymd_hm("2019-11-11T12:34"),
  "2019-11",          ymd_hm("2019-11-11T12:34"),
  "2019-12-04",       ymd_hm("2019-11-11T12:34")
) |>
  derive_vars_dtm(
    dtc                = AESTDTC,
    new_vars_prefix    = "AST",
    highest_imputation = "M",
    date_imputation    = "first",
    time_imputation    = "first",
    min_dates          = exprs(TRTSDTM)
  )

ae3
```

### Avoid Imputed Dates After Data Cut-off or Death

```{r max-dates}
ae4 <- tribble(
  ~AEENDTC,           ~DTHDT,           ~DCUTDT,
  "2019-08-09T12:34", ymd("2019-11-11"), ymd("2019-12-02"),
  "2019-11",          ymd("2019-11-11"), ymd("2019-12-02"),
  "2019-12",          NA,                ymd("2019-12-02"),
  "2019-12-04",       NA,                ymd("2019-12-02")
) |>
  derive_vars_dtm(
    dtc                = AEENDTC,
    new_vars_prefix    = "AEN",
    highest_imputation = "M",
    date_imputation    = "last",
    time_imputation    = "last",
    max_dates          = exprs(DTHDT, DCUTDT)
  )

ae4
```

### Using Multiple Imputation Rules via `slice_derivation()`

```{r slice-derivation}
vs <- tribble(
  ~VSDTC,              ~VSTPT,
  "2019-08-09T12:34", NA,
  "2019-10-12",       "PRE-DOSE",
  "2019-11-10",       NA,
  "2019-12-04",       NA
) |>
  slice_derivation(
    derivation = derive_vars_dtm,
    args = params(
      dtc             = VSDTC,
      new_vars_prefix = "A"
    ),
    derivation_slice(
      filter = VSTPT == "PRE-DOSE",
      args   = params(time_imputation = "first")
    ),
    derivation_slice(
      filter = TRUE,
      args   = params(time_imputation = "last")
    )
  )

vs
```

## See Also

-   [Creating ADSL](adsl.qmd)
-   [Creating an OCCDS ADaM](occds.qmd)
-   [Higher Order Functions](higher_order.qmd)
