# Introduction to admiral

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Main Idea

The main idea of `{admiral}` is that an ADaM dataset is built by a sequence of
**derivations**. Each derivation adds one or more variables or records to the
processed dataset. This modular approach makes it easy to adjust code by adding,
removing, or modifying derivations. Each derivation is a function call.

In this chapter we will explore some of the different types of derivation
functions offered by `{admiral}`, as well as the argument conventions they
follow and how best to start an `{admiral}` script.

## Setup

To work through the examples below we need a few packages and some example
data.

```{r setup-packages}
#| message: false
#| warning: false
library(dplyr)
library(lubridate)
library(stringr)
library(tibble)
library(pharmaversesdtm)
library(admiral)
```

```{r load-data}
# Read in SDTM datasets from pharmaversesdtm
dm <- pharmaversesdtm::dm
ds <- pharmaversesdtm::ds
ex <- pharmaversesdtm::ex
vs <- pharmaversesdtm::vs

# Use the admiral example ADSL
admiral_adsl <- admiral::admiral_adsl
```

The `adsl` and `advs` objects below are prepared to showcase
[addition of variables](#add-variables) and
[addition of records](#add-records) later.

```{r prepare-data}
ex_ext <- ex |>
  derive_vars_dtm(
    dtc = EXSTDTC,
    new_vars_prefix = "EXST"
  )

vs <- vs |>
  filter(
    USUBJID %in% c(
      "01-701-1015", "01-701-1023", "01-703-1086",
      "01-703-1096", "01-707-1037", "01-716-1024"
    ) &
      VSTESTCD %in% c("SYSBP", "DIABP") &
      VSPOS == "SUPINE"
  )

adsl <- admiral_adsl |>
  select(-TRTSDTM, -TRTSTMF)

advs <- vs |>
  mutate(
    PARAM    = VSTEST,
    PARAMCD  = VSTESTCD,
    AVAL     = VSSTRESN,
    AVALU    = VSORRESU,
    AVISIT   = VISIT,
    AVISITN  = VISITNUM
  )
```

::: callout-note
In the example above we read dummy R data from `{pharmaversesdtm}`. However, if
you are using SAS datasets as a starting point, be sure to consult
[Handling of Missing Values](concepts_conventions.qmd#missing) in the
[Programming Concepts and Conventions](concepts_conventions.qmd) chapter to learn
how and why you should use `convert_blanks_to_na()` during this process.
:::

## Derivation Functions

The most important functions in `{admiral}` are the
[derivations](https://pharmaverse.github.io/admiral/reference/index.html#derivations-for-adding-variables).
Derivations add variables or observations/records to the input dataset. Existing
variables and observations of the input dataset are **not** changed. Derivation
functions start with `derive_`. The first argument of these functions expects
the input dataset. This allows derivations to be chained together using the
native pipe `|>` (or `%>%`).

Functions which derive a dedicated variable start with `derive_var_` followed by
the variable name, e.g., `derive_var_trtdurd()` derives `TRTDURD`.

Functions which can derive multiple variables start with `derive_vars_` followed
by the variable names, e.g., `derive_vars_dtm()` can derive both `TRTSDTM`
and `TRTSTMF`.

Functions which derive a dedicated parameter start with `derive_param_` followed
by the parameter name, e.g., `derive_param_bmi()` derives the `BMI` parameter.

### Adding Variables {#add-variables}

Below is an example call to one of the most common derivation functions,
`derive_vars_merged()`. This function adds variable(s) to the input dataset
based on the contents of another dataset. In this example, we add the treatment
start datetime and corresponding imputation flag (`TRTSTMF`) to `adsl` by
identifying the first record in `ex` with a non-missing Exposure Start Datetime
(`EXSTDTM`) when sorting by `EXSTDTM` and `EXSEQ`.

```{r add-variables}
adsl <- adsl |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )
```

```{r show-add-variables, echo=FALSE}
adsl |>
  select(USUBJID, TRTSDTM, TRTSTMF) |>
  head(5)
```

### Adding Records {#add-records}

Another common derivation function is `derive_param_computed()`. This function
adds a derived parameter to an input dataset. In the example below, we use it
to derive Mean Arterial Pressure (MAP) from Systolic and Diastolic blood
pressure values. The parameters needed for the derivation are specified in
`parameters`, and within `set_values_to` we set all the variable values for the
new derived record.

```{r add-records}
advs <- advs |>
  derive_param_computed(
    by_vars       = exprs(USUBJID, AVISIT, AVISITN),
    parameters    = c("SYSBP", "DIABP"),
    set_values_to = exprs(
      AVAL    = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3,
      PARAMCD = "MAP",
      PARAM   = "Mean Arterial Pressure (mmHg)",
      AVALU   = "mmHg"
    )
  )
```

```{r show-add-records, echo=FALSE}
advs |>
  arrange(USUBJID, AVISITN, PARAMCD) |>
  select(USUBJID, AVISIT, PARAMCD, AVAL) |>
  head(10)
```

::: callout-tip
For the users' convenience, `{admiral}` actually provides `derive_param_map()`
(a wrapper of `derive_param_computed()`) to derive MAP. The above example
serves for illustrative purposes only.
:::

## Other Types of Functions

Along with derivation functions, `{admiral}` provides a large collection of
helper functions to support ADaM derivations. Here are some of the other
categories.

### Higher Order Functions

[Higher order functions](higher_order.qmd) are `{admiral}` functions that take
other functions as input. They enhance the existing portfolio of derivation
functions by allowing greater customisation of the latter's behaviour. A
derivation function can be:

-   `call_derivation()` — Called multiple times, while varying some of the
    input arguments.
-   `restrict_derivation()` — Executed on a subset of the input dataset.
-   `slice_derivation()` — Executed differently on subsets of the input
    dataset.

Higher order functions are a relatively advanced topic within `{admiral}`. You
can read all about them in the dedicated [Higher Order Functions](higher_order.qmd)
chapter.

### Computation Functions

[Computations](https://pharmaverse.github.io/admiral/reference/index.html#computation-functions-for-vectors)
expect vectors as input and return a vector. These functions cannot be used with
the pipe operator directly. They can be used in expressions like
`convert_dtc_to_dt()` in the derivation of the Final Lab Visit Date
(`FINLABDT`) in the example below:

```{r computation-example}
# Add the date of the final lab visit to ADSL
adsl_finlab <- dm |>
  derive_vars_merged(
    dataset_add = ds,
    by_vars     = exprs(USUBJID),
    new_vars    = exprs(FINLABDT = convert_dtc_to_dt(DSSTDTC)),
    filter_add  = DSDECOD == "FINAL LAB VISIT"
  )
```

```{r show-computation, echo=FALSE}
adsl_finlab |>
  select(STUDYID, USUBJID, FINLABDT) |>
  head(5)
```

Computations can also be used inside a `mutate()` statement:

```{r mutate-example}
adsl_finlab <- adsl_finlab |>
  mutate(RFSTDT = convert_dtc_to_dt(RFSTDTC))
```

```{r show-mutate, echo=FALSE}
adsl_finlab |>
  select(STUDYID, USUBJID, RFSTDTC, RFSTDT) |>
  head(5)
```

### Filter Functions

[Filter functions](https://pharmaverse.github.io/admiral/reference/index.html#utilities-for-filtering-observations)
filter the input dataset in different ways, for instance returning records that
fit or do not fit a certain condition, or that are the first/last observation
in a by-group. These functions form an important internal backbone to some of
the `{admiral}` functions, but can also be used on their own to explore or
manipulate a dataset. For instance, in the example below we use
`filter_extreme()` to extract the most recent MAP records in `advs`:

```{r filter-example}
advs_lastmap <- advs |>
  filter(PARAMCD == "MAP") |>
  filter_extreme(
    by_vars = exprs(USUBJID),
    order   = exprs(AVISITN, PARAMCD),
    mode    = "last"
  )
```

```{r show-filter, echo=FALSE}
advs_lastmap |>
  select(USUBJID, AVISIT, PARAMCD, AVAL)
```

## Argument Conventions

Within the `{admiral}` package, any arguments which expect variable names or
expressions of variable names must be specified as **symbols** or
**expressions** rather than strings.

-   For arguments which expect a **single variable name**, the name can be
    specified without quotes, e.g. `new_var = TEMPBL`.
-   For arguments which expect **one or more variable names**, a list of
    symbols is expected, e.g. `by_vars = exprs(PARAMCD, AVISIT)`.
-   For arguments which expect a **single expression**, the expression is
    passed as-is, e.g. `filter = PARAMCD == "TEMP"`.
-   For arguments which expect **one or more expressions**, a list of
    expressions is expected, e.g. `order = exprs(AVISIT, desc(AESEV))`.

If you are new to expressions, consider reading the
[Expressions in Scripts](concepts_conventions.qmd#exprs) section of the
[Programming Concepts and Conventions](concepts_conventions.qmd) chapter to
learn more.

## Starting a Script

For each ADaM data structure, the following chapters provide an overview of the
workflow and example function calls for the most common derivation steps:

-   [Creating ADSL](adsl.qmd)
-   [Creating an OCCDS ADaM](occds.qmd)
-   [Creating a BDS Findings ADaM](bds_finding.qmd)

`{admiral}` also provides template R scripts as a starting point. They can be
created by calling `use_ad_template()`, e.g.:

```r
use_ad_template(
  adam_name = "adsl",
  save_path  = "./ad_adsl.R"
)
```

A list of all available templates can be obtained by `list_all_templates()`:

```{r list-templates}
list_all_templates()
```

## Getting Help

Support is provided via the
[admiral Slack channel](https://pharmaverse.slack.com/). Additionally,
please feel free to raise issues in the
[GitHub repository](https://github.com/pharmaverse/admiral/issues).

## See Also

-   [Template scripts](https://github.com/pharmaverse/admiral/tree/main/inst/templates)
-   [Programming Concepts and Conventions](concepts_conventions.qmd)
-   [Programming Strategy](https://pharmaverse.github.io/admiraldev/articles/programming_strategy.html)
