# Lab Grading

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Lab grading assigns a numeric or character severity grade to laboratory
results based on reference ranges or pre-defined criteria. Common
grading schemes include:

-   **CTCAE** — Common Terminology Criteria for Adverse Events
    (FDA/NCI-defined; most widely used in oncology).
-   **Division of AIDS (DAIDS)** grading criteria.
-   **WHO** grading criteria.

The `{admiral}` package provides the `derive_var_atoxgr_dir()` and
`derive_var_atoxgr()` functions to apply grading criteria, along with
built-in lookup tables for CTCAE.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
```

## Programming Workflow

1.  [Prepare the Lab Dataset](#prepare)
2.  [Apply Grading Criteria (Low and High)](#grading)
3.  [Derive the Overall Toxicity Grade](#overall)
4.  [Baseline Grade](#baseline)
5.  [Maximum Grade](#max)

## Prepare the Lab Dataset {#prepare}

```{r prepare}
lb   <- pharmaversesdtm::lb   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl

adlb <- lb |>
  filter(USUBJID %in% unique(adsl$USUBJID)[1:6]) |>
  mutate(
    PARAMCD = LBTESTCD,
    PARAM   = LBTEST,
    AVAL    = LBSTRESN,
    AVALU   = LBSTRESU,
    ANRLO   = LBSTNRLO,
    ANRHI   = LBSTNRHI,
    A1LO    = LBSTNRLO,  # lower reference range (used for grading)
    A1HI    = LBSTNRHI   # upper reference range (used for grading)
  )

adlb |>
  select(USUBJID, PARAMCD, AVAL, AVALU, ANRLO, ANRHI) |>
  head(8)
```

## Apply Grading Criteria {#grading}

The `derive_var_atoxgr_dir()` function derives the toxicity grade for a
given direction (low or high). It uses a look-up table of grading
criteria.

The `{admiral}` package provides `atoxgr_criteria_ctcv5` (CTCAE v5),
`atoxgr_criteria_ctcv4` (CTCAE v4), and `atoxgr_criteria_daids` (DAIDS).

```{r grading}
# Derive toxicity grade for values below the normal range (low)
adlb <- adlb |>
  derive_var_atoxgr_dir(
    new_var      = ATOXGRL,
    tox_grade_crit_var = PARAMCD,
    aval_var     = AVAL,
    atoxgr_criteria = atoxgr_criteria_ctcv5,
    direction    = "L"
  )

# Derive toxicity grade for values above the normal range (high)
adlb <- adlb |>
  derive_var_atoxgr_dir(
    new_var      = ATOXGRH,
    tox_grade_crit_var = PARAMCD,
    aval_var     = AVAL,
    atoxgr_criteria = atoxgr_criteria_ctcv5,
    direction    = "H"
  )

adlb |>
  select(USUBJID, PARAMCD, AVAL, ANRLO, ANRHI, ATOXGRL, ATOXGRH) |>
  filter(!is.na(ATOXGRL) | !is.na(ATOXGRH)) |>
  head(8)
```

## Derive the Overall Toxicity Grade {#overall}

The `derive_var_atoxgr()` function combines the low and high grades
into a single overall grade. By convention:

-   Positive integers (1–4) represent high (above normal) grades.
-   Negative integers (−1 to −4) represent low (below normal) grades.

```{r overall}
adlb <- adlb |>
  derive_var_atoxgr()

adlb |>
  select(USUBJID, PARAMCD, AVAL, ANRLO, ANRHI, ATOXGRL, ATOXGRH, ATOXGR) |>
  filter(!is.na(ATOXGRL) | !is.na(ATOXGRH)) |>
  head(8)
```

## Baseline Grade {#baseline}

Baseline grade (`BTOXGR`) is derived by flagging the baseline record
first (`ABLFL`) and then deriving the baseline value:

```{r baseline}
adlb <- adlb |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(new_vars_prefix = "A", dtc = LBDTC) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = ATOXGR,
    new_var    = BTOXGR
  )

adlb |>
  select(USUBJID, PARAMCD, ADT, AVAL, ABLFL, ATOXGR, BTOXGR) |>
  filter(!is.na(ATOXGR)) |>
  head(8)
```

## Maximum Grade {#max}

To derive the maximum toxicity grade for each subject and parameter,
use `derive_summary_records()` or `derive_var_extreme_flag()`:

```{r max-grade}
adlb_max <- adlb |>
  filter(!is.na(ATOXGR)) |>
  group_by(STUDYID, USUBJID, PARAMCD) |>
  summarise(ATOXGRMX = max(ATOXGR, na.rm = TRUE), .groups = "drop")

adlb_max |>
  filter(ATOXGRMX > 0) |>
  head(8)
```

## Grading Criteria Datasets

The following look-up datasets are provided by `{admiral}`:

| Dataset | Description |
|---|---|
| `atoxgr_criteria_ctcv5` | NCI CTCAE v5.0 criteria |
| `atoxgr_criteria_ctcv4` | NCI CTCAE v4.0 criteria |
| `atoxgr_criteria_daids` | DAIDS grading table (Nov 2014) |

Each dataset contains the grading criteria by parameter code (`PARAMCD`),
direction, and grade threshold. See `?atoxgr_criteria_ctcv5` for the
complete column specification.

## Example Script

An ADLB template that includes toxicity grading is available:

```r
use_ad_template("ADLB")
```

## See Also

-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [Hy's Law](hys_law.qmd)
