# Creating an OCCDS ADaM

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter describes creating an OCCDS (Occurrence Data Structure) ADaM.
Examples are presented and tested in the context of `ADAE`. However, the
examples could be applied to other OCCDS ADaMs such as `ADCM`, `ADMH`,
`ADDV`, etc.

::: callout-note
All examples assume CDISC SDTM and/or ADaM format as input unless otherwise
specified.
:::

## Programming Workflow

1.  [Read in Data](#readdata)
2.  [Derive/Impute End and Start Analysis Date/time and Relative Day](#datetime)
3.  [Derive Durations](#duration)
4.  [Derive Planned and Actual Treatment](#trtpa)
5.  [Derive Date/Date-time of Last Dose](#last_dose)
6.  [Derive Treatment Dose and Unit](#dose_unit)
7.  [Derive Severity, Causality, and Toxicity Grade](#severity)
8.  [Derive Treatment Emergent Flag](#trtflag)
9.  [Derive Occurrence Flags](#occflag)
10. [Derive Query Variables](#query)
11. [Add ADSL Variables](#adsl_vars)
12. [Derive Analysis Sequence Number](#aseq)
13. [Add Labels and Attributes](#attributes)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADAE` should be read into
the environment. This will be a company-specific process. Some of the data
frames needed may be `AE` and `ADSL`.

For example purposes, the CDISC Pilot SDTM and ADaM datasets — which are
included in `{pharmaversesdtm}` — are used.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
library(tibble)
```

```{r load-data}
ae   <- pharmaversesdtm::ae |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl
ex_single <- admiral::ex_single
```

```{r filter-data, echo=FALSE}
ae <- filter(ae, USUBJID %in% c(
  "01-701-1015", "01-701-1023", "01-703-1086",
  "01-703-1096", "01-707-1037", "01-716-1024"
))
```

At this step it may be useful to join `ADSL` to your `AE` domain as well. Only
the `ADSL` variables used for derivations are selected at this step. The rest
of the relevant `ADSL` variables are added later.

```{r adae-base}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, DTHDT, EOSDT)

adae <- derive_vars_merged(
  ae,
  dataset_add = adsl,
  new_vars    = adsl_vars,
  by          = exprs(STUDYID, USUBJID)
)

adae |>
  select(USUBJID, AESEQ, AETERM, AESTDTC, TRTSDT, TRTEDT, TRT01A, TRT01P, DTHDT, EOSDT) |>
  head(5)
```

## Derive/Impute End and Start Analysis Date/time and Relative Day {#datetime}

This section derives `ASTDTM`, `ASTDT`, `ASTDY`, `AENDTM`, `AENDT`, and
`AENDY`. The function `derive_vars_dtm()` can be used to derive `ASTDTM` and
`AENDTM`. `ASTDT` and `AENDT` can be derived from `ASTDTM` and `AENDTM`
respectively using `derive_vars_dtm_to_dt()`. `derive_vars_dy()` can be used
to create `ASTDY` and `AENDY`.

```{r datetime}
adae <- adae |>
  derive_vars_dtm(
    dtc                = AESTDTC,
    new_vars_prefix    = "AST",
    highest_imputation = "M",
    min_dates          = exprs(TRTSDT)
  ) |>
  derive_vars_dtm(
    dtc                = AEENDTC,
    new_vars_prefix    = "AEN",
    highest_imputation = "M",
    date_imputation    = "last",
    time_imputation    = "last",
    max_dates          = exprs(DTHDT, EOSDT)
  ) |>
  derive_vars_dtm_to_dt(exprs(ASTDTM, AENDTM)) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ASTDT, AENDT)
  )

adae |>
  select(USUBJID, AESTDTC, AEENDTC, ASTDTM, ASTDT, ASTDY, AENDTM, AENDT, AENDY) |>
  head(5)
```

See also [Date and Time Imputation](imputation.qmd).

## Derive Durations {#duration}

The function `derive_vars_duration()` can be used to create the variables
`ADURN` and `ADURU`.

```{r duration}
adae <- adae |>
  derive_vars_duration(
    new_var      = ADURN,
    new_var_unit = ADURU,
    start_date   = ASTDT,
    end_date     = AENDT
  )

adae |>
  select(USUBJID, AESTDTC, AEENDTC, ASTDT, AENDT, ADURN, ADURU) |>
  head(5)
```

## Derive Planned and Actual Treatment {#trtpa}

`TRTA` and `TRTP` must match at least one value of the character treatment
variables in ADSL (e.g., `TRTxxA`/`TRTxxP`).

An example of a simple implementation for a study without periods:

```{r trtpa}
adae <- mutate(adae, TRTP = TRT01P, TRTA = TRT01A)
count(adae, TRTP, TRTA, TRT01P, TRT01A)
```

For studies with periods, see [Visit and Period Variables](visits_periods.qmd#treatment_bds).

## Derive Date/Date-time of Last Dose {#last_dose}

The function `derive_vars_joined()` can be used to derive the last dose date
before the start of the event.

```{r last-dose}
ex_single <- derive_vars_dtm(
  ex_single,
  dtc             = EXSTDTC,
  new_vars_prefix = "EXST",
  flag_imputation = "none"
)

adae <- derive_vars_joined(
  adae,
  ex_single,
  by_vars     = exprs(STUDYID, USUBJID),
  new_vars    = exprs(LDOSEDTM = EXSTDTM),
  join_vars   = exprs(EXSTDTM),
  join_type   = "all",
  order       = exprs(EXSTDTM),
  filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & grepl("PLACEBO", EXTRT))) & !is.na(EXSTDTM),
  filter_join = EXSTDTM <= ASTDTM,
  mode        = "last"
)

adae |>
  select(USUBJID, AEDECOD, AESEQ, AESTDTC, AEENDTC, ASTDT, AENDT, LDOSEDTM) |>
  head(5)
```

## Derive Treatment Dose and Unit {#dose_unit}

In a similar manner, you could derive the treatment dose and unit at the time
of the event.

```{r dose-unit}
ex_single <- derive_vars_dtm(
  ex_single,
  dtc             = EXENDTC,
  new_vars_prefix = "EXEN",
  time_imputation = "last",
  flag_imputation = "none"
)

adae <- derive_vars_joined(
  adae,
  ex_single,
  by_vars     = exprs(STUDYID, USUBJID),
  new_vars    = exprs(DOSEON = EXDOSE, DOSEU = EXDOSU),
  join_vars   = exprs(EXSTDTM, EXENDTM),
  join_type   = "all",
  filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & grepl("PLACEBO", EXTRT))) & !is.na(EXSTDTM),
  filter_join = EXSTDTM <= ASTDTM & (ASTDTM <= EXENDTM | is.na(EXENDTM))
)

adae |>
  select(USUBJID, AEDECOD, AESEQ, AESTDTC, AEENDTC, ASTDT, AENDT, DOSEON, DOSEU) |>
  head(5)
```

## Derive Severity, Causality, and Toxicity Grade {#severity}

The variables `ASEV`, `AREL`, and `ATOXGR` can be added using simple
`dplyr::mutate()` assignments if no imputation is required.

```{r severity}
adae <- adae |>
  mutate(
    ASEV = AESEV,
    AREL = AEREL
  )
```

## Derive Treatment Emergent Flag {#trtflag}

To derive the treatment emergent flag `TRTEMFL`, use `derive_var_trtemfl()`.
In the example below, we use a 30-day window after treatment end.

```{r trtemfl}
adae <- adae |>
  derive_var_trtemfl(
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 30
  )

adae |>
  select(USUBJID, TRTSDT, TRTEDT, AESTDTC, ASTDT, TRTEMFL) |>
  head(5)
```

## Derive Occurrence Flags {#occflag}

The function `derive_var_extreme_flag()` can help derive variables such as
`AOCCIFL`, `AOCCPIFL`, `AOCCSIFL`, and `AOCCzzFL`.

Flag first occurrence of most severe adverse event:

```{r occflag}
adae <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(USUBJID),
      order   = exprs(
        as.integer(factor(
          ASEV,
          levels = c("DEATH THREATENING", "SEVERE", "MODERATE", "MILD")
        )),
        ASTDTM, AESEQ
      ),
      new_var = AOCCIFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

adae |>
  select(USUBJID, ASTDTM, ASEV, AESEQ, TRTEMFL, AOCCIFL) |>
  head(5)
```

## Derive Query Variables {#query}

For deriving query variables such as `SMQzzNAM`, `SMQzzCD`, `SMQzzSC`,
`SMQzzSCN`, or `CQzzNAM`, the `derive_vars_query()` function can be used.
See [Queries Dataset](queries_dataset.qmd) for a detailed description.

```{r query}
queries <- admiral::queries

adae1 <- tribble(
  ~USUBJID, ~ASTDTM,               ~AETERM,                               ~AESEQ, ~AEDECOD,                              ~AELLT,       ~AELLTCD,
  "01",     "2020-06-02 23:59:59", "ALANINE AMINOTRANSFERASE ABNORMAL",   3,      "Alanine aminotransferase abnormal",   NA_character_, NA_integer_,
  "02",     "2020-06-05 23:59:59", "BASEDOW'S DISEASE",                   5,      "Basedow's disease",                   NA_character_, 1L,
  "03",     "2020-06-07 23:59:59", "SOME TERM",                           2,      "Some query",                          "Some term",   NA_integer_,
  "05",     "2020-06-09 23:59:59", "ALVEOLAR PROTEINOSIS",                7,      "Alveolar proteinosis",                NA_character_, NA_integer_
)

adae_query <- derive_vars_query(dataset = adae1, dataset_queries = queries)
adae_query |> head(5)
```

## Add ADSL Variables {#adsl_vars}

If needed, the other `ADSL` variables can now be added:

```{r adsl-vars}
adae <- adae |>
  derive_vars_merged(
    dataset_add = select(adsl, !!!negate_vars(adsl_vars)),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adae |>
  select(USUBJID, AEDECOD, ASTDTM, DTHDT, RFSTDTC, RFENDTC, AGE, AGEU, SEX) |>
  head(5)
```

## Derive Analysis Sequence Number {#aseq}

The function `derive_var_obs_number()` can be used for deriving `ASEQ` to
ensure uniqueness of subject records within the dataset.

```{r aseq}
adcm <- tribble(
  ~USUBJID,       ~ASTDTM,          ~CMSEQ, ~CMDECOD,         ~ATC1CD, ~ATC2CD, ~ATC3CD, ~ATC4CD,
  "BP40257-1001", "2013-07-05 UTC", "14",   "PARACETAMOL",    "N",     "N02",   "N02B",  "N02BE",
  "BP40257-1001", "2013-08-15 UTC", "18",   "SOLUMEDROL",     "D",     "D10",   "D10A",  "D10AA",
  "BP40257-1001", "2013-08-15 UTC", "18",   "SOLUMEDROL",     "D",     "D07",   "D07A",  "D07AA",
  "BP40257-1001", "2013-08-15 UTC", "18",   "SOLUMEDROL",     "H",     "H02",   "H02A",  "H02AB",
  "BP40257-1002", "2012-12-15 UTC", "19",   "SPIRONOLACTONE", "C",     "C03",   "C03D",  "C03DA"
)

adcm |>
  derive_var_obs_number(
    by_vars    = exprs(USUBJID),
    order      = exprs(ASTDTM, CMSEQ, ATC1CD, ATC2CD, ATC3CD, ATC4CD),
    new_var    = ASEQ,
    check_type = "error"
  )
```

## Add Labels and Attributes {#attributes}

See the equivalent section in [Creating ADSL](adsl.qmd#attributes) for a
discussion of packages like `{metacore}`, `{metatools}`, and `{xportr}` for
applying metadata.

## Example Script

An ADAE template script is available in the `{admiral}` package:

```r
use_ad_template("ADAE")
```

## See Also

-   [Date/Time Imputation](imputation.qmd)
-   [Queries Dataset](queries_dataset.qmd)
-   [Visit and Period Variables](visits_periods.qmd)
