# PK Non-Compartmental Analysis (ADNCA)

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This chapter describes creating a PK Non-Compartmental Analysis ADaM
dataset (`ADNCA`). NCA analyses require concentration-time data from
the PC (Pharmacokinetic Concentrations) SDTM domain and dosing
information from the EX domain.

::: callout-note
NCA analyses are often performed using specialist PK software (e.g.
Phoenix WinNonlin, R `{PKNCA}`). This chapter focuses on the ADaM
programming steps needed to prepare the input dataset.
:::

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(pharmaversesdtm)
library(lubridate)
```

## Programming Workflow

1.  [Read in Data](#readdata)
2.  [Derive Nominal and Actual Time Variables](#time)
3.  [Derive Dose Variables](#dose)
4.  [Assign `PARAMCD`, `PARAM`, `AVAL`](#paramcd)
5.  [Derive Relative Time Variables (`NFRLT`, `AFRLT`, `NRRLT`, `ARRLT`)](#rlt)
6.  [Add ADSL Variables](#adsl_vars)
7.  [Derive Analysis Sequence Number](#aseq)

## Read in Data {#readdata}

```{r load-data}
pc   <- pharmaversesdtm::pc   |> convert_blanks_to_na()
ex   <- pharmaversesdtm::ex   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl

# Filter to subjects with PK data
pk_subjects <- unique(pc$USUBJID)

pc   <- pc   |> filter(USUBJID %in% pk_subjects)
ex   <- ex   |> filter(USUBJID %in% pk_subjects)
adsl <- adsl |> filter(USUBJID %in% pk_subjects)
```

## Derive Nominal and Actual Time Variables {#time}

Actual sample collection datetime (`ADTM`) is derived from `PCDTC`. The
nominal time (`NFRLT`) comes from `PCTPTNUM` in the PC domain:

```{r time-vars}
adpc <- pc |>
  derive_vars_dtm(
    dtc             = PCDTC,
    new_vars_prefix = "A",
    flag_imputation = "none"
  ) |>
  mutate(
    NFRLT = PCTPTNUM,  # Nominal time from first dose (hours)
    ATPT  = PCTPT,
    ATPTN = PCTPTNUM
  )

adpc |>
  select(USUBJID, PCTRT, PCDTC, ADTM, NFRLT, ATPT) |>
  head(8)
```

## Derive Dose Variables {#dose}

Dosing records from the EX domain provide the reference dose event for
each concentration record:

```{r dose-vars}
ex_dtm <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    flag_imputation = "none"
  )

# Merge the most recent dose before the concentration time
adpc <- adpc |>
  derive_vars_joined(
    ex_dtm,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(EXSTDTM, EXDOSE, EXDOSU, EXTRT),
    join_vars   = exprs(EXSTDTM),
    join_type   = "all",
    filter_add  = EXDOSE > 0 & !is.na(EXSTDTM),
    filter_join = EXSTDTM <= ADTM,
    order       = exprs(EXSTDTM),
    mode        = "last"
  )

adpc |>
  select(USUBJID, PCTRT, ADTM, EXSTDTM, EXDOSE, EXTRT) |>
  head(8)
```

## Assign `PARAMCD`, `PARAM`, `AVAL` {#paramcd}

```{r paramcd}
adpc <- adpc |>
  mutate(
    PARAMCD = PCTESTCD,
    PARAM   = PCTEST,
    AVAL    = PCSTRESN,
    AVALU   = PCSTRESU,
    DTYPE   = NA_character_  # will be set for derived records
  )

adpc |>
  select(USUBJID, PARAMCD, PARAM, AVAL, AVALU) |>
  head(5)
```

## Derive Relative Time Variables {#rlt}

The key timing variables for NCA are:

-   `NFRLT` — Nominal time from first dose (hours)
-   `AFRLT` — Actual time from first dose (hours)
-   `NRRLT` — Nominal time from most recent dose (hours)
-   `ARRLT` — Actual time from most recent dose (hours)

```{r rlt}
adpc <- adpc |>
  mutate(
    AFRLT = as.numeric(difftime(ADTM, EXSTDTM, units = "hours")),
    NRRLT = NFRLT,   # In single-period studies, NRRLT = NFRLT
    ARRLT = AFRLT    # In single-period studies, ARRLT = AFRLT
  )

adpc |>
  select(USUBJID, ADTM, EXSTDTM, NFRLT, AFRLT, NRRLT, ARRLT) |>
  head(8)
```

## Add ADSL Variables {#adsl_vars}

```{r adsl-vars}
adpc <- adpc |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRT01A, TRT01P, TRTSDT, AGE, SEX, RACE, WEIGHTBL),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adpc |>
  select(USUBJID, PARAMCD, AVAL, TRT01A, AGE, SEX) |>
  head(5)
```

## Derive Analysis Sequence Number {#aseq}

```{r aseq}
adpc <- adpc |>
  derive_var_obs_number(
    by_vars    = exprs(STUDYID, USUBJID),
    order      = exprs(PARAMCD, ADTM, NFRLT),
    new_var    = ASEQ,
    check_type = "error"
  )

adpc |>
  select(USUBJID, PARAMCD, ADTM, NFRLT, ASEQ) |>
  head(5)
```

## Example Script

An ADNCA template script is available in the `{admiral}` package:

```r
use_ad_template("ADNCA")
```

## See Also

-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [Date/Time Imputation](imputation.qmd)
-   [Visit and Period Variables](visits_periods.qmd)
