# Queries Dataset

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

A **queries dataset** is a reference dataset that defines groupings of
preferred terms (PTs) or lower-level terms (LLTs) from MedDRA or
drug groupings. These groupings are used to derive additional variables
in OCCDS ADaM datasets such as `ADAE` and `ADCM`.

Examples of query types:

-   **Standardized MedDRA Queries (SMQs)** — groupings of MedDRA PTs and/or
    LLTs that identify adverse events related to a medical condition or
    area of special interest.
-   **Customised Queries (CQs)** — study- or company-specific groupings.
-   **Standardized Drug Groupings (SDGs)** — groupings of drugs by class
    or indication.

`{admiral}` provides the `create_query_data()` function to build a
queries dataset, and `derive_vars_query()` to derive the corresponding
variables in an ADaM dataset.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(tibble)
```

## Structure of the Queries Dataset

The queries dataset provided by `{admiral}` (`admiral::queries`) has the
following key columns:

| Column | Description |
|---|---|
| `PREFIX` | Prefix used in the derived variable names (e.g. `"SMQ01"`) |
| `GRPNAME` | Name of the query group |
| `GRPID` | Numeric ID of the group |
| `SCOPE` | Scope of the query (`"BROAD"` or `"NARROW"`) |
| `SCOPEN` | Numeric scope (1 = broad, 2 = narrow) |
| `SRCVAR` | Source variable in the ADaM dataset (`"AEDECOD"`, `"AELLT"`, etc.) |
| `TERMCHAR` | Character term value to match |
| `TERMNUM` | Numeric term code to match (`AELLTCD`) |

```{r show-queries}
admiral::queries |>
  select(PREFIX, GRPNAME, GRPID, SCOPE, SCOPEN, SRCVAR, TERMCHAR, TERMNUM) |>
  head(10)
```

## Using `derive_vars_query()`

`derive_vars_query()` adds query-related variables to an ADaM dataset.
For each row in the queries dataset, it looks for a match in the
specified `SRCVAR` column of the input dataset and, where found, adds
the corresponding group name and scope variables.

```{r derive-vars-query}
adae1 <- tribble(
  ~USUBJID, ~ASTDTM,               ~AETERM,                              ~AESEQ, ~AEDECOD,                             ~AELLT,        ~AELLTCD,
  "01",     "2020-06-02 23:59:59", "ALANINE AMINOTRANSFERASE ABNORMAL",  3,      "Alanine aminotransferase abnormal",  NA_character_, NA_integer_,
  "02",     "2020-06-05 23:59:59", "BASEDOW'S DISEASE",                  5,      "Basedow's disease",                  NA_character_, 1L,
  "03",     "2020-06-07 23:59:59", "SOME TERM",                          2,      "Some query",                         "Some term",   NA_integer_,
  "05",     "2020-06-09 23:59:59", "ALVEOLAR PROTEINOSIS",               7,      "Alveolar proteinosis",               NA_character_, NA_integer_
)

adae_with_queries <- derive_vars_query(
  dataset         = adae1,
  dataset_queries = admiral::queries
)

adae_with_queries |>
  select(USUBJID, AEDECOD, starts_with("SMQ"), starts_with("CQ")) |>
  head(5)
```

## Deriving SDG Variables for ADCM

Similarly, `derive_vars_query()` can be used to derive Standardized Drug
Grouping (SDG) variables:

```{r sdg}
sdg <- tribble(
  ~PREFIX, ~GRPNAME,          ~GRPID, ~SCOPE,  ~SCOPEN, ~SRCVAR,   ~TERMCHAR,          ~TERMNUM,
  "SDG01", "Diuretics",           11, "BROAD", 1,       "CMDECOD", "Diuretic 1",       NA,
  "SDG01", "Diuretics",           11, "BROAD", 1,       "CMDECOD", "Diuretic 2",       NA,
  "SDG02", "Corticosteroids",     12, "BROAD", 1,       "CMDECOD", "Corticosteroid 1", NA,
  "SDG02", "Corticosteroids",     12, "BROAD", 1,       "CMDECOD", "Corticosteroid 2", NA,
  "SDG02", "Corticosteroids",     12, "BROAD", 1,       "CMDECOD", "Corticosteroid 3", NA,
)

adcm <- tribble(
  ~USUBJID, ~ASTDTM,               ~CMDECOD,
  "01",     "2020-06-02 23:59:59", "Diuretic 1",
  "02",     "2020-06-05 23:59:59", "Diuretic 1",
  "03",     "2020-06-07 23:59:59", "Corticosteroid 2",
  "05",     "2020-06-09 23:59:59", "Diuretic 2"
)

derive_vars_query(adcm, sdg)
```

## Creating a Queries Dataset with `create_query_data()`

`create_query_data()` builds a queries dataset from a structured list of
`query()` objects. This is useful when SMQ or SDG definitions are
managed programmatically (e.g. sourced from a validated look-up table):

```{r create-query-data}
my_smq <- query(
  prefix  = "SMQ01",
  name    = "My Custom Query",
  id      = 42,
  scope   = "BROAD",
  add_scope_num = TRUE,
  definition = list(
    basket_select(
      name    = "My Custom Query",
      scope   = "BROAD",
      type    = "smq"
    )
  )
)
```

::: callout-note
The `basket_select()` function retrieves SMQ/SDG definitions from a
validated MedDRA dictionary. In a real implementation, this step would
be configured to connect to your organisation's MedDRA database.
:::

## See Also

-   [Creating an OCCDS ADaM](occds.qmd)
-   [Generic Derivations](generic.qmd)
