# Questionnaires

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Many clinical trials collect patient-reported outcomes (PROs) and
clinical outcome assessments (COAs) via questionnaires. The responses
are stored in SDTM QS (Questionnaire) datasets and are used to derive
composite scores in the corresponding ADaM dataset (e.g. `ADQS`).

`{admiral}` provides `derive_summary_records()` for aggregating item-level
responses into composite scores.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(dplyr)
library(tibble)
library(pharmaversesdtm)
library(lubridate)
```

## Typical Questionnaire Workflow

1.  [Read in QS data and define parameters](#params)
2.  [Derive item-level analysis values](#item-aval)
3.  [Derive composite scores](#composite)
4.  [Derive timing, baseline, and change from baseline](#timing)
5.  [Add ADSL variables](#adsl-vars)

## Read in QS Data and Define Parameters {#params}

```{r load-data}
qs   <- pharmaversesdtm::qs   |> convert_blanks_to_na()
adsl <- admiral::admiral_adsl

# Select a subset of subjects and questionnaire items for illustration
qs_sub <- qs |>
  filter(
    USUBJID %in% unique(adsl$USUBJID)[1:4],
    QSCAT == "ALZHEIMER'S DISEASE ASSESSMENT SCALE"
  )

qs_sub |>
  select(USUBJID, QSTESTCD, QSTEST, QSSTRESN, QSDTC, VISIT) |>
  head(10)
```

## Derive Item-Level Analysis Values {#item-aval}

Each QS item is mapped to `PARAMCD`, `PARAM`, and `AVAL`:

```{r item-aval}
adqs <- qs_sub |>
  mutate(
    PARAMCD = QSTESTCD,
    PARAM   = QSTEST,
    AVAL    = QSSTRESN,
    AVALC   = QSSTRESC
  ) |>
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc             = QSDTC
  ) |>
  mutate(
    AVISIT  = VISIT,
    AVISITN = VISITNUM
  )

adqs |>
  select(USUBJID, PARAMCD, PARAM, AVAL, ADT, AVISIT) |>
  head(8)
```

## Derive Composite Scores {#composite}

A composite score (e.g. total score) can be derived by summing or
averaging item-level values using `derive_summary_records()`:

```{r composite}
# Example: total ADAS-Cog score as the sum of individual items
adqs_total <- adqs |>
  derive_summary_records(
    by_vars       = exprs(STUDYID, USUBJID, VISIT, VISITNUM, ADT),
    filter_add    = str_detect(PARAMCD, "^ACITM"),  # ADAS-Cog items only
    set_values_to = exprs(
      AVAL    = sum(AVAL, na.rm = FALSE),   # NA if any item is missing
      PARAMCD = "ACTOT",
      PARAM   = "ADAS-Cog Total Score"
    )
  )

adqs_total |>
  filter(PARAMCD == "ACTOT") |>
  select(USUBJID, PARAMCD, PARAM, AVAL, VISIT) |>
  head(5)
```

::: callout-warning
Using `na.rm = FALSE` in the aggregation ensures that the total score is
`NA` if any item is missing, which is the recommended approach for most
questionnaire totals. Consult your SAP for the specific rule.
:::

## Derive Baseline and Change from Baseline {#timing}

```{r baseline}
adqs_total <- adqs_total |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRTSDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  ) |>
  derive_var_chg() |>
  derive_var_pchg()

adqs_total |>
  filter(PARAMCD == "ACTOT") |>
  select(USUBJID, VISIT, AVAL, ABLFL, BASE, CHG, PCHG) |>
  head(8)
```

## Add ADSL Variables {#adsl-vars}

```{r adsl-vars}
adqs_total <- adqs_total |>
  derive_vars_merged(
    dataset_add = adsl,
    new_vars    = exprs(TRT01A, TRT01P, AGE, SEX),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adqs_total |>
  select(USUBJID, PARAMCD, AVAL, TRT01A, AGE, SEX) |>
  filter(!is.na(AVAL)) |>
  head(5)
```

## Example Script

A QS template script is available in the `{admiral}` package:

```r
use_ad_template("ADQS")
```

## See Also

-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [Generic Derivations](generic.qmd)
-   [Higher Order Functions](higher_order.qmd)
