# Visit and Period Variables

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

The derivation of visit variables like `AVISIT`, `AVISITN`, `AWLO`,
`AWHI`, … or period, subperiod, or phase variables like `APERIOD`,
`TRT01A`, `TRT02A`, `ASPER`, `PHSDTM`, `PHEDTM`, … is highly
study-specific. Therefore `{admiral}` cannot provide functions that derive
these variables automatically. However, for common scenarios like visit
assignments based on time windows or deriving BDS period variables from
ADSL period variables, functions are provided that support those
derivations.

```{r load-packages}
#| message: false
#| warning: false
library(admiral)
library(tibble)
library(dplyr)
library(lubridate)
```

## Visit Variables (`AVISIT`, `AVISITN`, `AWLO`, `AWHI`, …) {#visits}

The most common ways of deriving `AVISIT` and `AVISITN` are:

1.  The variables are set to the collected visits (`VISIT` and
    `VISITNUM`) — achieved simply by calling `mutate()`.
2.  The variables are set based on time windows.

For the time-window approach, a study-specific reference dataset is
needed that provides for each visit the start and end day (`AWLO` and
`AWHI`) and the values of other visit-related variables.

```{r visit-windows}
windows <- tribble(
  ~AVISIT,    ~AWLO, ~AWHI, ~AVISITN, ~AWTARGET,
  "BASELINE",   -30,     1,        0,         1,
  "WEEK 1",       2,     7,        1,         5,
  "WEEK 2",       8,    15,        2,        11,
  "WEEK 3",      16,    22,        3,        19,
  "WEEK 4",      23,    30,        4,        26
)
```

Then the visits can be assigned based on analysis day (`ADY`) by calling
`derive_vars_joined()`:

```{r visit-assignment}
adbds <- tribble(
  ~USUBJID, ~ADY,
  "1",       -33,
  "1",        -2,
  "1",         3,
  "1",        24,
  "2",        NA,
)

adbds |>
  derive_vars_joined(
    dataset_add = windows,
    filter_join = AWLO <= ADY & ADY <= AWHI,
    join_type   = "all",
  )
```

## Period, Subperiod, and Phase Variables

If periods, subperiods, or phases are used, the corresponding variables
have to be consistent across all datasets. This can be achieved by
defining the periods, subperiods, or phases once and then using this
definition for all datasets, stored either in ADSL or in a separate
period/phase reference dataset.

### Creating the Period Reference Dataset from EX

```{r period-ref-from-ex}
ex <- tribble(
  ~STUDYID, ~USUBJID, ~VISIT,    ~EXTRT,   ~EXSTDTC,
  "xyz",    "1",      "Day 1",   "Drug X", "2022-01-02",
  "xyz",    "1",      "Week 4",  "Drug X", "2022-02-05",
  "xyz",    "1",      "Week 16", "Drug Y", "2022-05-03",
  "xyz",    "1",      "Week 28", "Drug Y", "2022-08-04",
  "xyz",    "2",      "Day 1",   "Drug Y", "2023-10-20",
  "xyz",    "2",      "Week 16", "Drug X", "2024-02-20",
  "xyz",    "2",      "Week 28", "Drug X", "2024-05-21"
)

adsl <- tribble(
  ~STUDYID, ~USUBJID, ~TRTSDT,           ~TRTEDT,           ~EOSDT,
  "xyz",    "1",      ymd("2022-01-02"), ymd("2022-08-04"), ymd("2022-09-10"),
  "xyz",    "2",      ymd("2023-10-20"), ymd("2024-05-21"), ymd("2024-06-30")
)

period_ref <- ex |>
  filter(VISIT %in% c("Day 1", "Week 16")) |>
  mutate(
    APERIOD  = case_when(VISIT == "Day 1" ~ 1, VISIT == "Week 16" ~ 2),
    TRTA     = EXTRT,
    APERSDT  = convert_dtc_to_dt(EXSTDTC)
  ) |>
  arrange(USUBJID, APERSDT) |>
  group_by(USUBJID) |>
  mutate(APEREDT = lead(APERSDT) - 1) |>
  ungroup() |>
  select(-starts_with("EX"), -VISIT) |>
  left_join(adsl, by = c("STUDYID", "USUBJID")) |>
  mutate(APEREDT = if_else(APERIOD == 2, EOSDT, APEREDT)) |>
  select(-EOSDT, -TRTSDT, -TRTEDT)

period_ref
```

### Creating the Phase Reference Dataset from ADSL {#from_adsl}

```{r phase-ref-from-adsl}
adsl1 <- adsl |>
  mutate(
    PH1SDT  = TRTSDT,
    PH1EDT  = TRTEDT + 28,
    APHASE1 = "TREATMENT",
    PH2SDT  = TRTEDT + 29,
    PH2EDT  = EOSDT,
    APHASE2 = "FUP"
  )

phase_ref <- create_period_dataset(
  adsl1,
  new_vars = exprs(PHSDT = PHwSDT, PHEDT = PHwEDT, APHASE = APHASEw)
)

phase_ref
```

## Creating ADSL Period Variables {#periods_adsl}

```{r adsl-period-vars}
adsl2 <- derive_vars_period(
  adsl,
  dataset_ref = period_ref,
  new_vars    = exprs(APxxSDT = APERSDT, APxxEDT = APEREDT)
)

adsl2 |>
  select(STUDYID, USUBJID, AP01SDT, AP01EDT, AP02SDT, AP02EDT)
```

## Creating BDS and OCCDS Period Variables {#periods_bds}

```{r bds-period-vars}
adae <- tribble(
  ~STUDYID, ~USUBJID, ~ASTDT,
  "xyz",    "1",      "2022-01-31",
  "xyz",    "1",      "2022-05-02",
  "xyz",    "1",      "2022-09-03",
  "xyz",    "2",      "2023-12-25",
  "xyz",    "2",      "2024-06-19",
) |>
  mutate(ASTDT = ymd(ASTDT))

adae |>
  derive_vars_joined(
    dataset_add = phase_ref,
    by_vars     = exprs(STUDYID, USUBJID),
    filter_join = PHSDT <= ASTDT & ASTDT <= PHEDT,
    join_type   = "all"
  )
```

## Treatment Variables (`TRTxxP`, `TRTxxA`, …)

### Creating ADSL Treatment Variables {#treatment_adsl}

```{r adsl-trt}
adsl3 <- derive_vars_period(
  adsl,
  dataset_ref = period_ref,
  new_vars    = exprs(APxxSDT = APERSDT, APxxEDT = APEREDT, TRTxxA = TRTA)
)

adsl3 |>
  select(STUDYID, USUBJID, TRT01A, TRT02A, AP01SDT, AP01EDT, AP02SDT, AP02EDT)
```

### Creating BDS and OCCDS Treatment Variables {#treatment_bds}

```{r bds-trt}
adae2 <- tribble(
  ~STUDYID, ~USUBJID, ~ASTDT,
  "xyz",    "1",      "2022-01-31",
  "xyz",    "1",      "2022-05-02",
  "xyz",    "2",      "2023-12-25",
  "xyz",    "2",      "2024-06-07",
) |>
  mutate(ASTDT = ymd(ASTDT)) |>
  derive_vars_joined(
    dataset_add = period_ref,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(APERIOD, TRTA),
    join_vars   = exprs(APERSDT, APEREDT),
    join_type   = "all",
    filter_join = APERSDT <= ASTDT & ASTDT <= APEREDT
  )

adae2
```

If period/phase variables exist in ADSL, `create_period_dataset()` can
create a reference dataset from them:

```{r period-from-adsl-vars}
period_ref2 <- adsl3 |>
  create_period_dataset(
    new_vars = exprs(APERSDT = APxxSDT, APEREDT = APxxEDT, TRTA = TRTxxA)
  )

period_ref2
```

## See Also

-   [Creating ADSL](adsl.qmd)
-   [Creating a BDS Findings ADaM](bds_finding.qmd)
-   [Creating an OCCDS ADaM](occds.qmd)
